---
title: "Figure 6 Notebook: Statistical Analysis of Public Mtb Genomes and generation of Oxidative Stress Response (OSR) Network"
output: html_notebook
---

**Author: Evan Pepper-Tunick**

Last updated: Nov 19, 2025

This notebook is for analyzing the variants identified in the Mtb clinical isolate dataset using a Bayesian probability framework and temporal analysis of Mtb evolution in the context of drug resistance mutations. This dataset includes isolates from: Silcocks et al. 2023

WGS data was processed and analyzed using an adaptation of the variant calling pipeline from Liu et al. 2022, which can be found at: https://github.com/baliga-lab/bwa_pipeline. Details on specific pipeline parameters for calling mutations and use of TB-Profiler can be found in the Methods section.

Two important files analyzed here are not included in the public distribution of this work. **Both of these files will be needed to run the code in this notebook.** The files are as follows:
1. Phenotypic drug susceptibility profiles of Mtb strains analyzed in this work. This data can be made available upon request.
2. CRISPRi strains during various drug treatments from S. Li et. al. 2022. This Source Data Fig. 1 from this publication and can be downloaded at: 10.1038/s41564-022-01130-y

**Important note: to run notebook, user's computer must have ≥ 10.0 GB memory**

```{r}
# importing necessary libraries
suppressMessages(
  {
  library(ggplot2)
  library(ggridges)
  library(ggupset)
  library(tidyverse)
  library(purrr)
  library(drc)
  library(knitr)
  library(factoextra)
  library(naniar)
  library(rstatix)
  library(ggpubr)
  library(scales)
  library(ggrepel)
  library(see)
  library(viridis)
  library(seqinr)
  library(Biostrings)
  }
)

# setting plot theme attributes
plot.theme <- theme(title = element_text(size = 14),
                    axis.text = element_text(size = 12),
                    axis.title = element_text(size = 12),
                    legend.title = element_text(size = 10),
                    legend.text = element_text(size = 8),
                    strip.text = element_text(size = 10))
```

```{r}
#################################################################################
# writing a function to calculate bayes probability
calculateBayes <- function(total.n, total.n.with.a, total.n.with.b, total.n.with.a.and.b) {
  # Calculate the conditional probability P(A|B)
  bayes.prob <- total.n.with.a.and.b / total.n.with.b
  return(bayes.prob)
}
# function for using apply to get bayes probs for each locus
calculateMetrics <- function(count.table, phenotypes) {
  # Get the sample IDs with mutations (non-zero counts)
  isomut <- colnames(count.table)[apply(count.table, 2, function(x) any(x > 0))]
  isolate.with.mutation.n <- length(isomut)
  # Get the count of resistant isolates with mutations
  R.isolate.n.with.mutation <- sum(resistant_isolates %in% isomut)
  # Calculate the Bayes probability
  bayes.prob <- calculateBayes(isolate.n, R.isolate.n, isolate.with.mutation.n, R.isolate.n.with.mutation)
  return(bayes.prob)
}
# function for converting matrix of isolate IDs and mutations to wide form
flattenCounts <- function(isolate.locus.matrix) {
  # transforming table to wide form with SNP occurances
  count.table.wide <- isolate.locus.matrix %>%
    group_by(Sample.ID, Locus) %>%
    summarise(n = n(), .groups = 'drop') %>%
    pivot_wider(names_from = Sample.ID, values_from = n, values_fill = list(n = 0))
  # convert to dataframe
  count.table.wide <- as.data.frame(count.table.wide)
  count.table.wide <- count.table.wide[complete.cases(count.table.wide), ]
  # set rownames and remove locus column
  rownames(count.table.wide) <- count.table.wide$Locus
  count.table.wide <- count.table.wide %>% dplyr::select(-Locus)
  return(count.table.wide)
}
#################################################################################
```

```{r}
# reading in metadata for clinical isolates with variants called
# THIS PHENOTYPES FILE CAN BE AVAILABLE UPON REQUEST
phenotypes <- read_csv(file = '../data-public-upon-request/dst-isolate-phenotypes.csv')
genotypes <- read_csv(gzfile('../data/Fig6/dst-isolates/dst-isolate-locus-matrix.csv.gz'))
names(genotypes) <- c('Sample.ID', 'Locus')
# insuring phenotype data only contains samples with variant data
phenotypes <- phenotypes %>% filter(Sample.ID %in% unique(genotypes$Sample.ID))
mtb.gene.map <- read_csv(file = '../data/genome-annotations/mtb-h37rv-full-genome-annotation.csv', show_col_types = FALSE)
gene.name.conversion.table <- read_csv(file = '../data/genome-annotations/gene-locus-map.csv')

# removing outlier from phylogeny analysis
outlier.id <- phenotypes %>% filter(major.lineage == 8) %>% pull(Sample.ID)
phenotypes <- phenotypes %>% filter(Sample.ID != outlier.id)
genotypes <- genotypes %>% filter(Sample.ID != outlier.id)
```

```{r}
# getting summary of lineages and resistance profiles for plotting
# generating bar plot of sizes of each lineage
phenotypes$major.lineage <- factor(phenotypes$major.lineage, levels = c(1, 2, 4))
lineages.summarized <- phenotypes %>% group_by(major.lineage) %>% tally() %>% filter(!is.na(major.lineage))
```

```{r}
# lineage counts
lineage.counts <- ggplot(lineages.summarized, aes(x = major.lineage, y = n, fill = major.lineage)) +
  geom_bar(stat = "identity", color = 'black') +
  geom_text(aes(label = n), nudge_y = 0.25, size = 5) +
  scale_y_log10() +
  xlab('Major Lineage') + ylab('Sample Count') +
  theme_minimal() +
  plot.theme +
  theme(legend.position = 'none')

pdf(file = '../figures/figure-panels/fig6-panel-a.pdf', height = 4, width = 3)
print(lineage.counts)
dev.off()
```

**Figure 6 Panel A**
```{r}
lineage.counts
```

```{r}
# preparing data from genotype calls to make summary figure for mutations in dataset
full.genotype.calls <- read_csv(gzfile('../data/Fig6/dst-isolates/dst-isolate-genotypes-all-pipelines-filtered.csv.gz'))
# full.genotype.calls <- read_csv(gzfile('../data/Fig6/dst-isolates/dst-isolate-genotypes.csv.gz'))
# cleaning up effect column for summarizing
full.genotype.calls.effect.change <- full.genotype.calls %>%
  filter(EFFECT %in% c('Intergenic', 'Nonsynonymous')) %>%
  mutate(EFFECT = case_when(EFFECT == 'Intergenic' ~ 'INTERGENIC',
                            EFFECT == 'Nonsynonymous' ~ 'NONSYNONYMOUS'))

full.genotypes <- rbind(full.genotype.calls %>% filter(!EFFECT %in% c('Intergenic', 'Nonsynonymous', 'EXON',
                                                                      'GENE_FUSION', 'GENE_FUSION_REVERESE')),
                        full.genotype.calls.effect.change)
full.genotypes <- full.genotypes %>%
  mutate(EFFECT = case_when(EFFECT == 'CODON_CHANGE_PLUS_CODON_DELETION' ~ 'INFRAME\nDELETION',
                            EFFECT == 'CODON_CHANGE_PLUS_CODON_INSERTION' ~ 'INFRAME\nINSERTION',
                            EFFECT == 'CODON_DELETION' ~ 'INFRAME\nDELETION',
                            EFFECT == 'CODON_INSERTION' ~ 'INFRAME\nINSERTION',
                            EFFECT == 'FRAME_SHIFT' ~ 'FRAME SHIFT',
                            EFFECT == 'NON_SYNONYMOUS_CODING' ~ 'MISSENSE',
                            EFFECT == 'NONSYNONYMOUS' ~ 'MISSENSE',
                            EFFECT == 'INTERGENIC' ~ 'INTERGENIC',
                            EFFECT == 'START_LOST' ~ 'START LOST',
                            EFFECT == 'STOP_GAINED' ~ 'STOP GAINED',
                            EFFECT == 'STOP_LOST' ~ 'STOP LOST'
                            ))
# tallying up calls for manuscript
n.fixed <- nrow(full.genotypes %>% filter(FREQ >= 90))
n.unfixed <- nrow(full.genotypes %>% filter(FREQ < 90))

n.nonsyn <- nrow(full.genotypes %>% filter(EFFECT != 'INTERGENIC'))
n.intergenic <- nrow(full.genotypes %>% filter(EFFECT == 'INTERGENIC'))
# to identify XX fixed and YY unfixed (at least 10% alternative allele frequency) mutations (XX intergenic and YY nonsynonymous protein-coding mutations) across 5,233 genomic loci within each sample
print(paste('We identified ', n.fixed, ' fixed mutations and ', n.unfixed, ' unfixed mutations', sep = ''))
print(paste('Across all variants, ', n.intergenic, ' were intergenic and ', n.nonsyn, ' were nonsynonymous', sep = ''))
```

```{r}
# creating figure 6 panel c
# summary of mutations in dataset
variant.summary.data <- full.genotypes %>%
  group_by(EFFECT) %>% tally() %>% arrange(desc(n))
variant.summary.data <- variant.summary.data %>%
  mutate(EFFECT = fct_reorder(EFFECT, -n))

variant.summary <- ggplot(variant.summary.data, aes(x = EFFECT, y = n)) +
  geom_bar(stat = "identity", color = 'black', fill = 'NA') +
  geom_text(aes(label = n), nudge_y = 0.25, size = 4) +
  scale_y_log10() +
  xlab('') + ylab('Variant Count') +
  theme_minimal() +
  plot.theme +
  theme(legend.position = 'none',
        axis.text.x = element_text(angle = 45, size = 8))

pdf(file = '../figures/figure-panels/fig6-panel-c.pdf', height = 4.4, width = 5)
print(variant.summary)
dev.off()
```


**Figure 6 Panel C**
```{r}
variant.summary
```

```{r}
# preparing data for upset plot for RIPE (RIF, INH, PZA, EMB)
# selecting columns of interest
status.data <- phenotypes %>%
  select(Sample.ID, Rifampicin, Isoniazid, Ethambutol, Streptomycin)
# making sure all values are either 1 or 0 (no NAs)
status.data <- replace(status.data, is.na(status.data), 0)
# creating vector of unique sample names
sample.ids <- status.data$Sample.ID
# creating a vector of all antibiotic names
abxs <- colnames(status.data[, 2:ncol(status.data)])
# for loop for getting abx name for all row values = 1 for each sample
sample.spectrums <- list()
for (i in 1:length(status.data$Sample.ID)) {
  spectrum.list <- list()
  sample <- status.data[i, 2:ncol(status.data)]
  if (sum(sample) == 0) {
    spectrum.list[i] <- NA
  }
  else {
    for (k in 1:length(abxs)) {
      if (sample[k] == 1) {
        spectrum.list[k] <- abxs[k]
      }
    }
  }
  
  sample.spectrums[[i]] <- unlist(spectrum.list)
}
# add list of resistance spectrum for each isolate to data as a new column
status.data$spectrum <- sample.spectrums
```

```{r}
# making upset plot
status.data.filtered <- status.data %>% distinct(Sample.ID, .keep_all = TRUE)
status.data.tally <- status.data.filtered %>% group_by(spectrum) %>% tally() %>% arrange(desc(n))
status.data.tally$spectrum <- unlist(lapply(status.data.tally$spectrum, function(x) ifelse(is.null(x), NA, x)))

write_csv(status.data.tally, file = '../data/Fig6/susceptibility-profile-counts.csv')

upset.plot <- ggplot(status.data.filtered, aes(x = spectrum)) +
  geom_bar(fill = 'tomato4', color = 'black') +
  scale_x_upset(n_intersections = 8) +
  scale_y_log10() +
  xlab('') + ylab('') +
  theme_minimal() +
  theme_combmatrix(combmatrix.label.text = element_text(size = 12),
                   combmatrix.panel.point.color.fill = "tomato3",
                   combmatrix.panel.point.size = 5,
                   combmatrix.panel.line.size = 0) +
  plot.theme

pdf(file = '../figures/figure-panels/fig6-panel-b.pdf', height = 3.75, width = 6)
print(upset.plot)
dev.off()
```


**Figure 4 Panel B**
```{r}
upset.plot
```

```{r}
# creating mutation counts for multi-locus sites
# each for loop checks for a unique combination of loci using 'and' / 'or' logic
genotype.isolates <- unique(genotypes$Sample.ID)
# mshA or mshC
sample.id.calls.1 <- list()
for (g in 1:length(genotype.isolates)) {
  isolate.data <- genotypes %>% filter(Sample.ID == genotype.isolates[g])
  if ('mshA' %in% isolate.data$Locus | 'mshC' %in% isolate.data$Locus) {
     sample.id.calls.1[[g]] <- tibble(Sample.ID = isolate.data$Sample.ID[1], Locus = 'mshA|mshC')
  }
  else {
    next
  }
}
extra.genotype.calls.1 <- do.call(rbind, sample.id.calls.1)
# mshA or mshB or mshC or mshD
sample.id.calls.2 <- list()
for (g in 1:length(genotype.isolates)) {
  isolate.data <- genotypes %>% filter(Sample.ID == genotype.isolates[g])
  if ('mshA' %in% isolate.data$Locus | 'mshB' %in% isolate.data$Locus | 'mshC' %in% isolate.data$Locus | 'mshD' %in% isolate.data$Locus) {
    sample.id.calls.2[[g]] <- tibble(Sample.ID = isolate.data$Sample.ID[1], Locus = 'mshA|mshB|mshC|mshD')
  }
  else {
    next
  }
}
extra.genotype.calls.2 <- do.call(rbind, sample.id.calls.2)

# binding to existing genotype data
genotypes <- rbind(genotypes, extra.genotype.calls.1, extra.genotype.calls.2) %>% arrange(Sample.ID)
```

```{r}
# calculating bayes probabilities and using output from NullBayesGlobal.R to read in null distributions for p value calculation
isolate.n <- nrow(phenotypes)
#################################################################################
# flattening SNP data to prepare for calculations
print('widening data (global)')
system.time(snp.data.wide <- flattenCounts(genotypes))
print('data widened (global)')
#################################################################################
# setting major values
# Get the sample IDs of isolates that are resistant to Isoniazid
resistant_isolates <- phenotypes %>%
  filter(Sample.ID %in% colnames(snp.data.wide) & Isoniazid == 1) %>%
  pull(Sample.ID)
R.isolate.n <- length(resistant_isolates)
#################################################################################
print('computing bayes probabilities (global)')
system.time(bayes <- data.frame(sapply(1:nrow(snp.data.wide), function(x){ calculateMetrics(snp.data.wide[x, ], phenotypes) })))
print('done computing bayes probabilities (global)')
names(bayes) <- 'bayes'
bayes$locus <- rownames(snp.data.wide)
bayes <- bayes %>% arrange(desc(bayes))
write_csv(bayes, file = '../data/Fig6/LDF/bayes-probabilities/INH-Global.csv')
#################################################################################
```

```{r}
bayes <- read_csv(file = '../data/Fig6/LDF/bayes-probabilities/INH-Global.csv', show_col_types = FALSE)
path.to.nulls <- '../data/Fig6/LDF/bayes-null-distributions/INH-global'
# aggregating null distribution data
null.files <- list.files(path = path.to.nulls)
null.dists <- list()
print('reading in nulls (global)')
# for (i in 1:5000) {
for (i in 1:length(null.files)) {
  null.dists[[i]] <- read_csv(file = paste(path.to.nulls, '/', null.files[i], sep = ''), show_col_types = FALSE)
}
null.bayes <- do.call(rbind, null.dists)
print('done reading in all nulls (global)')
```

```{r}
# visualizing null distributions
distributions <- ggplot(bayes, aes(x = bayes)) +
  geom_histogram(binwidth = 0.004, fill = 'red', alpha = 0.8) +
  geom_density(data = null.bayes, aes(x = bayes), fill = 'black', alpha = 0.4) +
  geom_vline(xintercept = min(null.bayes$bayes), linetype = 'dashed') +
  geom_vline(xintercept = max(null.bayes$bayes), linetype = 'dashed') +
  geom_vline(xintercept = mean(null.bayes$bayes)) +
  annotate(geom = 'text', x = (mean(null.bayes$bayes) + 0.05), y = (length(null.files) / 25),
           label = paste('mean\n', round(mean(null.bayes$bayes), digits = 4), sep = '')) +
  xlab('Bayes Probabilities') + ylab('') +
  ggtitle(paste('Bayes Probabilities for INH resistance (red)\nNull probability distribution, Global DST isolates,\n', length(null.files), ' iterations (black)', sep = '')) +
  theme_minimal() +
  plot.theme
```

```{r}
distributions
```


```{r}
# generating p values for bayes probabilities
p.values <- list()
for (b in 1:nrow(bayes)) {
  locus <- bayes[b, ]
  bayes.value <- locus$bayes
  # getting p value by comparing bayes probability to the null distribution
  # how many values in the null are smaller than the calculated bayes probability
  n.null.less.than <- length(which(null.bayes$bayes < bayes.value))
  # case when bayes probability is less than all null values
  if (n.null.less.than == 0) {
    p.value <- 1
  }
  else {
    # p value is equal to 1 minus the proportion of all random bayes values less than computed bayes value 
    p.value <- 1 - ( n.null.less.than / nrow(null.bayes) )
  }
  p.values[b] <- p.value
}
# adding p values back to data
bayes$p.value <- unlist(p.values)
bayes$adj.p.value <- p.adjust(bayes$p.value, method = 'fdr')
write_csv(bayes, file = '../data/Fig6/LDF/bayes-probabilities/INH-Global-PValues.csv')

info.table <- tibble(lineage = 'Global',
                     n.isolates = isolate.n,
                     n.inhR.isolates = R.isolate.n,
                     prop.inhR = (R.isolate.n / isolate.n),
                     mean.null = mean(null.bayes$bayes),
                     n.significant.loci = nrow(bayes %>% filter(adj.p.value <= 0.01)))
```

Reading in calculated bayes probabilities with p values

```{r}
bayes <- read_csv(file = '../data/Fig6/LDF/bayes-probabilities/INH-Global-PValues.csv', show_col_types = FALSE)
```

Creating part of Figure 6D

```{r}
ridges <- ggplot(bayes, aes(x = bayes)) +
  geom_density(alpha = 0.2, fill = 'black', linewidth = 0.25) +
  #scale_fill_manual(values = c('black', '#f37066', '#c9942c', '#76ac42', '#14b360')) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0), breaks = scales::pretty_breaks(n = 10)) +
  theme_minimal() +
  xlab('') + ylab('') +
  theme(legend.position = 'none')

pdf(file = '../figures/figure-panels/fig6-panel-d.pdf', height = 3.75, width = 6)
print(ridges)
dev.off()
```


**Figure 6 Panel D**
```{r}
ridges
```

```{r}
# Creating density plot for null distribution to overlay
null.bayes.sample <- sample_n(null.bayes, 100000, replace = F)
significance.threshold <- tail(bayes %>% filter(adj.p.value <= 0.05), n = 1) %>% pull(bayes)

null <- ggplot(null.bayes.sample, aes(x = bayes)) +
  geom_density(linewidth = 0.5, linetype = 'dashed', color = 'gray50', fill = 'NA') +
  geom_vline(xintercept = info.table$mean.null, linetype = 'dashed', color = 'black', alpha = 0.5) +
  geom_vline(xintercept = significance.threshold, linetype = 'dashed', color = 'red', alpha = 0.5) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0), breaks = scales::pretty_breaks(n = 10), limits = c(0, 1)) +
  theme_minimal() +
  xlab('') + ylab('') +
  theme(legend.position = 'none',
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 0),
        axis.text.y = element_text(size = 0))

pdf(file = '../figures/figure-panels/fig6-panel-d-null.pdf', height = 2, width = 6)
print(null)
dev.off()
```

Assessing Bayes probabilities amongst loci known or likely to be associated INH resistance.
Loci included in analysis are from: 10.1128/microbiolspec.mgm2-0014-2013

```{r}
# creating list of INH resistance associated loci
inhR.loci.list <- c('oxyR-ahpC', 'Rv2427A-Rv2428',
                    'ahpC', 'Rv2428',
                    'oxyR', 'Rv2427A',
                    'furA', 'Rv1909c',
                    'katG', 'Rv1908c',
                    'furA-katG', 'katG-furA', 'Rv1909c-Rv1908c', 'Rv1908c-Rv1909c',
                    'fabG-inhA', 'inhA-fabG', 'Rv1483-Rv1484', 'Rv1484-Rv1483',
                    'inhA', 'Rv1484',
                    'kasA', 'Rv2245',
                    'ndh', 'Rv1854c',
                    'ndhA', 'Rv0392c',
                    'iniB', 'Rv0341',
                    'iniA', 'Rv0342',
                    'iniC', 'Rv0343',
                    'Rv0340',
                    'nat', 'Rv3566c',
                    'Rv1592c',
                    'fadE24', 'Rv3139',
                    'Rv1772',
                    'efpA', 'Rv2846c',
                    'fabD', 'Rv2243',
                    'accD6', 'Rv2247',
                    'fbpC', 'Rv0129c')

bayes.inhR.loci <- bayes %>% filter(locus %in% inhR.loci.list)

```

Also assessing if there is an enrichment of intergenic loci amongst the significant Bayes probabilities

```{r}
# getting list of all loci with mutations to get population size and population successes
all.loci <- unique(bayes$locus)
n.total <- length(all.loci)
n.intergenic.total <- length(all.loci[grepl('-', all.loci)])
# getting sample size and sample successes
n.significant.loci <- length(unique(bayes %>% filter(adj.p.value <= 0.05) %>% pull(locus)))
n.significant.genic.loci <- length(unique(bayes[!grepl('-', bayes$locus), ] %>% filter(adj.p.value <= 0.05) %>% pull(locus)))
n.significant.intergenic.loci <- length(unique(bayes[grepl('-', bayes$locus), ] %>% filter(adj.p.value <= 0.05) %>% pull(locus)))

# performing hypergeometric test to calculate significance of enrichment of intergenic loci amonst significant hits
p.value <- phyper(q = (n.significant.intergenic.loci - 1),
                  m = n.intergenic.total,
                  n = ( n.total - n.intergenic.total ),
                  k = n.significant.loci,
                  lower.tail = FALSE)
print(paste('hypergeometric test p value for enrichment of intergenic loci amongst significant hits = ', p.value, sep = ''))
```

Next, performing analysis on overlap of genes enriched or depleted in the following genome-wide study:
S. Li et. al. 2022: https://doi.org/10.1038/s41564-022-01130-y (CRISPRi)

File "li-et-al-2022-table-SD-F1.xlsx" below is the Source Data Fig. 1 file from the above publication. Please download this file and save it as the same name to run this code.

```{r}
# performing hypergeometric tests to look for over-representation of significant BP genes enriched/depleted in CRISPRi data
# first converting gene names in bayes probabilities to Rv numbers
bayes.genic <- bayes[!grepl('-', bayes$locus), ]
bayes.genic <- left_join(bayes.genic, gene.name.conversion.table)
# importing CRISPRi gene enrichment/depletion data from Supplemental Data 2
treatments <- c('BDQ 1d', 'BDQ 5d', 'BDQ 10d', 'CLR 1d', 'CLR 5d', 'CLR 10d', 'EMB 1d', 'EMB 5d', 'EMB 10d', 'INH 1d', 'INH 5d', 'INH 10d', 'LVX 1d', 'LVX 5d', 'LVX 10d',
                'LZD 1d', 'LZD 5d', 'LZD 10d', 'RIF 1d', 'RIF 5d', 'RIF 10d', 'STR 1d', 'STR 5d', 'VAN 1d', 'VAN 5d', 'VAN 10d')
sheets <- c(3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17,
            18, 19, 20, 21, 22, 23, 24, 25, 27, 28, 29)

crispri.enrichment.test.list <- list()

for (d in 1:length(treatments)) {
  treatment <- treatments[d]
  selected.sheet <- sheets[d]
  
  crispri.data <- readxl::read_excel(path = '../data-public-upon-request/li-et-al-2022-table-SD-F1.xlsx', sheet = selected.sheet)
  
  crispri.data$`locus tag` <- paste('Rv', unlist(lapply(strsplit(unlist(lapply(strsplit(crispri.data$ORF_ID, ':'), '[', 1)), 'RVBD'), '[', 2)), sep = '')
  crispri.data[crispri.data == 'RvNA'] <- NA
  crispri.data$`gene name` <- unlist(lapply(strsplit(crispri.data$ORF_ID, ':'), '[', 2))
  crispri.data <- crispri.data %>% filter(`locus tag` %in% mtb.gene.map$locus)
  # high dose columns
  crispri.data.high <- crispri.data[, c(1, 22, 23, 26, 27, 28, 29)]
  names(crispri.data.high) <- c('ORF', 'neg l2fc', 'neg fdr', 'pos l2fc', 'pos fdr', 'locus tag', 'gene name')
  # med dose columns
  crispri.data.med <- crispri.data[, c(1, 14, 15, 18, 19, 28, 29)]
  names(crispri.data.med) <- c('ORF', 'neg l2fc', 'neg fdr', 'pos l2fc', 'pos fdr', 'locus tag', 'gene name')
  # low dose columns
  crispri.data.low <- crispri.data[, c(1, 6, 7, 10, 11, 28, 29)]
  names(crispri.data.low) <- c('ORF', 'neg l2fc', 'neg fdr', 'pos l2fc', 'pos fdr', 'locus tag', 'gene name')

  experiment.groups <- list(crispri.data.high, crispri.data.med, crispri.data.low)
  experiment.conds <- c('high', 'medium', 'low')
  crispri.hyper.tests.per.dose <- list()
  for (e in 1:length(experiment.groups)) {
    crispri.data.select <- experiment.groups[[e]]
    dose <- experiment.conds[e]
    # filtering bayes genes and enrichment/depletion data genes to only include genes that exist in both datasets
    overlapping.genes <- intersect(x = bayes.genic$rv.number, y = mtb.gene.map$locus)
    # filtering both datasets
    crispri.data.filtered <- crispri.data.select %>% filter(`locus tag` %in% unique(overlapping.genes))
    bayes.filtered <- bayes.genic %>% filter(rv.number %in% unique(crispri.data.filtered$`locus tag`))
    # getting population size for downstream hypergeometric tests
    population.size <- nrow(bayes.filtered)
    # getting mean and standard deviation of bayes distribution
    mean.bayes <- mean(bayes.filtered$bayes)
    sd.bayes <- sd(bayes.filtered$bayes)

    sig.genes <- bayes.filtered %>% filter(adj.p.value <= 0.05)
    # bayes.subset <- sig.genes[1:(nrow(sig.genes) * bayes.percentile), ]
    # getting input values for hypergeometric test
    # extracting list of significant genes (bayes)
    top.bayes.genes <- sig.genes$rv.number
    sample.size <- length(top.bayes.genes)
    #################################################################################
    # hypergeometric test for all significant genes from crispri
    # extracting list of significant genes (crispri)
    crispri.significant.genes <- unique(rbind(crispri.data.filtered %>%
                                                filter(`pos l2fc` >= 1, `pos fdr` <= 0.01),
                                              crispri.data.filtered %>%
                                                filter(`neg l2fc` <= -1, `neg fdr` <= 0.01)) %>% pull(`locus tag`))
    population.successes <- length(crispri.significant.genes)
    sample.successes <- length(intersect(x = top.bayes.genes, y = crispri.significant.genes))
    # saving results of hypergeometric test
    hyper.all <- tibble(
      dataset = 'CRISPRi',
      response = 'Responder',
      hyper.test.p.value = phyper(q = (sample.successes - 1),
                                  m = population.successes,
                                  n = ( population.size - population.successes ),
                                  k = sample.size,
                                  lower.tail = FALSE),
      population.size = population.size,
      n.significant.bayes.genes = sample.size,
      n.significant.exp.genes = population.successes, n.overlap = sample.successes,
      overlapping.genes = list(intersect(x = top.bayes.genes, y = crispri.significant.genes)))
    #################################################################################
    # hypergeometric test only for genes with alleviating or enriched interactions
    crispri.up.genes <- unique(crispri.data.filtered %>%
                                 filter(`pos l2fc` >= 1, `pos fdr` <= 0.01) %>% pull(`locus tag`))
    population.successes <- length(crispri.up.genes)
    sample.successes <- length(intersect(x = top.bayes.genes, y = crispri.up.genes))
    hyper.up <- tibble(
      dataset = 'CRISPRi',
      response = 'Enriched',
      hyper.test.p.value = phyper(q = (sample.successes - 1),
                                  m = population.successes,
                                  n = ( population.size - population.successes ),
                                  k = sample.size,
                                  lower.tail = FALSE),
      population.size = population.size,
      n.significant.bayes.genes = sample.size,
      n.significant.exp.genes = population.successes, n.overlap = sample.successes,
      overlapping.genes = list(intersect(x = top.bayes.genes, y = crispri.up.genes)))
    #################################################################################
    # hypergeometric test only for genes with aggravating or depleted interactions
    crispri.down.genes <- unique(crispri.data.filtered %>%
                                   filter(`neg l2fc` <= -1, `neg fdr` <= 0.01) %>% pull(`locus tag`))
    population.successes <- length(crispri.down.genes)
    sample.successes <- length(intersect(x = top.bayes.genes, y = crispri.down.genes))
    hyper.down <- tibble(
      dataset = 'CRISPRi',
      response = 'Depleted',
      hyper.test.p.value = phyper(q = (sample.successes - 1),
                                  m = population.successes,
                                  n = ( population.size - population.successes ),
                                  k = sample.size,
                                  lower.tail = FALSE),
      population.size = population.size,
      n.significant.bayes.genes = sample.size,
      n.significant.exp.genes = population.successes, n.overlap = sample.successes,
      overlapping.genes = list(intersect(x = top.bayes.genes, y = crispri.down.genes)))

    # aggregating all data from three hypergeometric tests
    hyper.tests <- rbind(hyper.all, hyper.up, hyper.down)
    hyper.tests$treatment <- paste(dose, ' dose ', treatment, sep = '')
    crispri.hyper.tests.per.dose[[e]] <- hyper.tests
  }
  # putting together all hypergeometric tests for CRISPRi
  crispri.hyper.tests <- do.call(rbind, crispri.hyper.tests.per.dose)
  crispri.enrichment.test.list[[d]] <- crispri.hyper.tests
}

crispri.enrichment.tests <- do.call(rbind, crispri.enrichment.test.list)
crispri.enrichment.tests.filtered <- crispri.enrichment.tests %>% filter(hyper.test.p.value <= 0.05)
```

```{r}
bayes.crispri.enrichment <- crispri.enrichment.tests.filtered
bayes.crispri.enrichment$overlapping.genes <- sapply(bayes.crispri.enrichment$overlapping.genes, function(x) paste(x, collapse = ', '))
write_csv(bayes.crispri.enrichment, file = '../data/Fig6/bayes-enrichment-in-crispri.csv')
```


```{r}
# creating volcano plot from Li et al. 2022, Figure 1C, to show overlap of genes with high Bayes probability
# this will also be Figure 6 Panel E
crispri.data <- readxl::read_excel(path = '../data-public-upon-request/li-et-al-2022-table-SD-F1.xlsx', sheet = 13)

crispri.data$`locus tag` <- paste('Rv', unlist(lapply(strsplit(unlist(lapply(strsplit(crispri.data$ORF_ID, ':'), '[', 1)), 'RVBD'), '[', 2)), sep = '')
crispri.data[crispri.data == 'RvNA'] <- NA
crispri.data$`gene name` <- unlist(lapply(strsplit(crispri.data$ORF_ID, ':'), '[', 2))
crispri.data <- crispri.data %>% filter(`locus tag` %in% mtb.gene.map$locus)
# high dose columns
crispri.data.high <- crispri.data[, c(1, 22, 23, 26, 27, 28, 29)]
names(crispri.data.high) <- c('ORF', 'neg l2fc', 'neg fdr', 'pos l2fc', 'pos fdr', 'locus tag', 'gene name')
# separating enriched and depleted data
pos.data <- crispri.data.high %>% dplyr::select(`pos l2fc`, `pos fdr`, `locus tag`, `gene name`)
pos.data$l2fc <- pos.data$`pos l2fc`
pos.data$log.10.fdr <- ( 1 * log10(pos.data$`pos fdr`) )
pos.data$Significance <- ifelse(pos.data$l2fc >= 1 & pos.data$log.10.fdr <= -2, 'Enriched', 'Not significant')
pos.data <- pos.data[!is.na(pos.data$Significance), ]
# repeating steps for depleted data
neg.data <- crispri.data.high %>% dplyr::select(`neg l2fc`, `neg fdr`, `locus tag`, `gene name`)
neg.data$l2fc <- neg.data$`neg l2fc`
neg.data$log.10.fdr <- ( 1 * log10(neg.data$`neg fdr`) )
neg.data$Significance <- ifelse(neg.data$l2fc <= -1 & neg.data$log.10.fdr <= -2, 'Depleted', 'Not significant')
neg.data <- neg.data[!is.na(neg.data$Significance), ]

# creating volcano plot
volcano <- ggplot() +
  geom_point(data = pos.data, aes(x = l2fc, y = log.10.fdr, color = Significance), alpha = 0.5) +
  geom_point(data = neg.data, aes(x = l2fc, y = log.10.fdr, color = Significance), alpha = 0.5) +
  geom_vline(xintercept = 1, linetype = 'dashed', color = 'grey30') +
  geom_vline(xintercept = -1, linetype = 'dashed', color = 'grey30') +
  geom_hline(yintercept = -2, linetype = 'dashed', color = 'grey30') +
  scale_color_manual(values = c('blue3', 'red3', 'grey')) +
  scale_x_continuous(position = "top") +
  xlab('L2FC') + ylab('log10 FDR') +
  theme_minimal() +
  plot.theme +
  theme(legend.position = c(0.85, 0.35),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

```{r}
# generating figure
pdf(file = '../figures/figure-panels/fig6-panel-e.pdf', height = 4, width = 4)
print(volcano)
dev.off()
```

**Figure 6 Panel E**
```{r}
volcano
```

Also performing analysis on GO terms:
antioxidant activity --> GO:0016209
cellular oxidant detoxification --> GO:0098869
removal of superoxide radicals --> GO:0019430
mycothiol-dependent detoxification --> GO:0010127
response to oxidative stress --> GO:0006979
response to nitrosative stress --> GO:0051409


```{r}
#################################################################################
go.1 <- readxl::read_excel(path = '../data/Fig6/oxidative-stress-network/oxidative-stress-genes.xlsx', sheet = 2)
go.2 <- readxl::read_excel(path = '../data/Fig6/oxidative-stress-network/oxidative-stress-genes.xlsx', sheet = 3)
go.3 <- readxl::read_excel(path = '../data/Fig6/oxidative-stress-network/oxidative-stress-genes.xlsx', sheet = 4)
go.4 <- readxl::read_excel(path = '../data/Fig6/oxidative-stress-network/oxidative-stress-genes.xlsx', sheet = 5)
go.5 <- readxl::read_excel(path = '../data/Fig6/oxidative-stress-network/oxidative-stress-genes.xlsx', sheet = 6)
go.6 <- readxl::read_excel(path = '../data/Fig6/oxidative-stress-network/oxidative-stress-genes.xlsx', sheet = 7)
go.data <- rbind(go.1, go.2, go.3, go.4, go.5, go.6)
go.data.filtered <- go.data %>% filter(`Gene Names (ordered locus)` %in% mtb.gene.map$locus)
go.data.filtered <- go.data.filtered %>% dplyr::select(`Gene Names (ordered locus)`, `GO Term`)
names(go.data.filtered) <- c('rv.number', 'GO Term')
oxidative.stress.genes <- go.data.filtered[match(unique(go.data.filtered$rv.number), go.data.filtered$rv.number), ]
oxidative.stress.genes <- left_join(oxidative.stress.genes, gene.name.conversion.table)
```

```{r}
# looking at TFOE data to extract regulons for oxidative stress network
tfoe.fc <- read_csv(file = '../data/Fig6/oxidative-stress-network/tfoe-fold-change.csv')
tfoe.fc <- tfoe.fc %>% pivot_longer(cols = -1, names_to = 'TF', values_to = 'FC')
tfoe.pval <- read_csv(file = '../data/Fig6/oxidative-stress-network/tfoe-p-values.csv')
tfoe.pval <- tfoe.pval %>% pivot_longer(cols = -1, names_to = 'TF', values_to = 'pvalue')
# column binding data
tfoe.data <- left_join(x = tfoe.fc, tfoe.pval)
tfoe.data$interaction <- ifelse(tfoe.data$FC < 0, 'repression', 'activation')
# reading in ChIP seq data for TF binding evidence
chip.seq <- read_csv(file = '../data/Fig6/oxidative-stress-network/chipseq-binding-evidence.csv')
chip.seq <- chip.seq %>% pivot_longer(cols = -1, names_to = 'TF', values_to = 'Evidence')
chip.seq.evidence <- na.omit(chip.seq)
# merging all tfoe and chipseq data
tfoe.chipseq <- left_join(tfoe.data, chip.seq)
tfoe.chipseq$adj.pvalue <- p.adjust(tfoe.chipseq$pvalue, method = 'fdr')
tfoe.chipseq$log10adj.pvalue <- log10(tfoe.chipseq$adj.pvalue)
# adding minimum pvalue to plot in log scale
tfoe.data$pvalue <- tfoe.data$pvalue + 7.46e-275
tfoe.data$adj.pvalue <- p.adjust(tfoe.data$pvalue, method = 'fdr')
tfoe.data$log10adj.pvalue <- log10(tfoe.data$adj.pvalue)

# printing out values for manuscript
print(paste('The RNAseq TFOE dataset contains information on ', length(unique(tfoe.data$TF)), ' TFs', sep = ''))
print(paste('The ChIP-seq TFOE dataset contains information on ', length(unique(chip.seq$TF)), ' TFs', sep = ''))
```

```{r}
# filtering data to only include significant interactions
tfoe.data.filtered <- tfoe.data %>% filter(adj.pvalue <= 0.05)
# setting pop. size parameter for hypergeometric test
population.size <- length(unique(tfoe.data.filtered$Gene))
# performing hypergeometric tests to identify which regulons are enriched for ROS genes
unique.tfs <- unique(tfoe.data.filtered$TF)
enrichment.data <- list()
for (t in 1:length(unique.tfs)) {
  # filter gene lists for each TF to get regulon
  regulon.data <- tfoe.data.filtered %>% filter(TF == unique.tfs[t])
  # setting remainder parameters for hypergeometric test
  regulated.genes <- unique(regulon.data$Gene)
  sample.size <- length(regulated.genes)
  population.successes <- nrow(oxidative.stress.genes)
  ros.genes <- oxidative.stress.genes$rv.number
  sample.successes <- length(intersect(x = regulated.genes, y = ros.genes))
  # saving results of hypergeometric test
  hyper.test <- tibble(
    TF = unique.tfs[t],
    n.regulated.genes = sample.size,
    n.ros.genes = population.successes, n.overlap = sample.successes,
    overlapping.loci = list(intersect(x = regulated.genes, y = ros.genes)),
    hyper.test.pvalue = phyper(q = (sample.successes - 1),
                               m = population.successes,
                               n = ( population.size - population.successes ),
                               k = sample.size,
                               lower.tail = FALSE))
  enrichment.data[[t]] <- hyper.test
}
network.enrichment <- do.call(rbind, enrichment.data)
network.enrichment$adj.pvalue <- p.adjust(network.enrichment$hyper.test.pvalue, method = 'fdr')
# filtering for significant hypergeometric tests to find regulons enriched for ROS genes
network.enrichment.filtered <- network.enrichment %>%
  filter(hyper.test.pvalue <= 0.05)
# capturing list of TFs with ROS-enriched regulons
ros.regulon.tfs <- unique(network.enrichment.filtered$TF)
# filter TFOE interactions for TFs identified from hypergeometric tests
tfoe.data.filtered.enriched <- tfoe.data.filtered %>% filter(TF %in% unique(network.enrichment.filtered$TF))
# prune network to only include ros genes
ros.network <- tfoe.data.filtered.enriched %>% filter(Gene %in% oxidative.stress.genes$rv.number)
# identifying the TFs with influence over the primary TFs of ROS genes
tfs.of.tfs <- tfoe.data.filtered %>% filter(Gene %in% unique(tfoe.data.filtered.enriched$TF))
# adding TFs of the TFs in the network back to the network itself
full.network <- rbind(ros.network, tfs.of.tfs) %>% arrange(TF)
# merging with ChIP-seq data to include interactions with evidence of binding
full.network.with.evidence <- left_join(full.network, chip.seq.evidence)
# adding in relationships from ChIPseq and TFOE evidence that was missed by enrichment analysis
# including all ROS gene interactions with evidence in both datasets
additional.data <- tfoe.chipseq %>% filter(Gene %in% oxidative.stress.genes$rv.number | TF %in% oxidative.stress.genes$rv.number,
                                                adj.pvalue <= 0.05, !is.na(Evidence))

# additional.data <- tfoe.chipseq %>% filter(Gene %in% oxidative.stress.genes$rv.number | TF %in% oxidative.stress.genes$rv.number,
#                                            !is.na(Evidence))

# removing instances of autoregulation
full.network.with.evidence <- full.network.with.evidence %>% filter(Gene != TF)
# merging with identified data
full.network.with.evidence <- rbind(full.network.with.evidence, additional.data)
full.network.with.evidence <- unique(full.network.with.evidence)
```

```{r}
# adding back into the network lone nodes of genes from the GO analysis that did not end up in the OSR network above
included.osr.genes <- intersect(x = c(full.network.with.evidence$Gene, full.network.with.evidence$TF), y = oxidative.stress.genes$rv.number)
extra.osr.nodes <- oxidative.stress.genes %>% filter(!rv.number %in% included.osr.genes)

# writing code to identify if specific osr gene is a Gene or TF based on inclusion in TFOE
tfoe.tfs <- unique(tfoe.data$TF)
extra.node.info.list <- list()
for (e in 1:nrow(extra.osr.nodes)) {
  selected.node <- extra.osr.nodes$rv.number[e]
  if (selected.node %in% tfoe.tfs) {
    extra.node.info.list[[e]] <- tibble(Gene = NA, TF = selected.node, FC = NA, pvalue = NA,
                                        interaction = NA, adj.pvalue = NA, log10adj.pvalue = NA, Evidence = NA)
  }
  else {
    extra.node.info.list[[e]] <- tibble(Gene = selected.node, TF = NA, FC = NA, pvalue = NA,
                                        interaction = NA, adj.pvalue = NA, log10adj.pvalue = NA, Evidence = NA)
  }
}
extra.node.info <- do.call(rbind, extra.node.info.list)
full.network.with.evidence <- rbind(full.network.with.evidence, extra.node.info)
```

```{r}
# annotating TFOE network
names(full.network.with.evidence)[1] <- 'rv.number'
full.network.with.evidence <- left_join(full.network.with.evidence, gene.name.conversion.table)
names(full.network.with.evidence)[c(1, 2, 9)] <- c('target.rv.number', 'rv.number', 'Target')
full.network.with.evidence <- left_join(full.network.with.evidence, gene.name.conversion.table)
names(full.network.with.evidence)[c(2, 10)] <- c('source.rv.number', 'Source')

full.network.with.evidence$Target <- ifelse(is.na(full.network.with.evidence$Target),
                                            full.network.with.evidence$target.rv.number,
                                            full.network.with.evidence$Target)

full.network.with.evidence$Source <- ifelse(is.na(full.network.with.evidence$Source),
                                            full.network.with.evidence$source.rv.number,
                                            full.network.with.evidence$Source)

# adding GO annotations in one column for either target, source, or both
annotations <- list()
for (r in 1:nrow(full.network.with.evidence)) {
  data <- full.network.with.evidence[r, ]
  if (data$target.rv.number %in% oxidative.stress.genes$rv.number & data$source.rv.number %in% oxidative.stress.genes$rv.number) {
    target.term <- oxidative.stress.genes %>% filter(rv.number == data$target.rv.number) %>% pull(`GO Term`)
    source.term <- oxidative.stress.genes %>% filter(rv.number == data$source.rv.number) %>% pull(`GO Term`)
    annotations[r] <- paste('Target: ', target.term, '; Source: ', source.term, sep = '')
  }
  else if (data$target.rv.number %in% oxidative.stress.genes$rv.number & !data$source.rv.number %in% oxidative.stress.genes$rv.number) {
    target.term <- oxidative.stress.genes %>% filter(rv.number == data$target.rv.number) %>% pull(`GO Term`)
    annotations[r] <- paste('Target: ', target.term, sep = '')
  }
  else if (!data$target.rv.number %in% oxidative.stress.genes$rv.number & data$source.rv.number %in% oxidative.stress.genes$rv.number) {
    source.term <- oxidative.stress.genes %>% filter(rv.number == data$source.rv.number) %>% pull(`GO Term`)
    annotations[r] <- paste('Source: ', source.term, sep = '')
  }
  else {
    annotations[r] <- 'No association'
  }
}

full.network.with.evidence$`GO annotation` <- unlist(annotations)

names(full.network.with.evidence) <- c('Target.rv.number', 'Source.rv.number', 'TFOE log2 FC', 'PValue', 'Interaction',
                                       'Adj.PValue', 'Adj.log10PValue', 'ChIP-seq Evidence', 'Target', 'Source',
                                       'GO annotation')
full.network.with.evidence <- full.network.with.evidence %>%
  dplyr::select(Source, Target, `TFOE log2 FC`, PValue, Adj.PValue, Interaction, `ChIP-seq Evidence`, `GO annotation`, Source.rv.number, Target.rv.number)
full.network.with.evidence$`Abs TFOE log2 FC` <- abs(full.network.with.evidence$`TFOE log2 FC`)

full.network.with.evidence <- unique(full.network.with.evidence)

# resplitting data between interaction network and lone OSR nodes
interaction.network <- full.network.with.evidence %>% filter(!is.na(Interaction))
lone.osr.nodes <- full.network.with.evidence %>% filter(is.na(Interaction))
```

```{r}
# writing to file
write_csv(interaction.network, file = '../data/Fig6/oxidative-stress-network/cytoscape-ros-network-input.csv')
write_csv(lone.osr.nodes, file = '../data/Fig6/oxidative-stress-network/cytoscape-ros-lone-nodes-input.csv')
```

```{r}
# analysis of genes from oxidative stress network and bayes
ros.network.bayes.genes <- bayes %>% filter(locus %in% full.network.with.evidence$Source | locus %in% full.network.with.evidence$Target,
                                            adj.p.value <= 0.05)
# looking separately at intergenic loci
intergenic.mapping <- gene.name.conversion.table
names(intergenic.mapping) <- c('full.locus', 'locus')
intergenic.mapping$upstream.gene <- unlist(lapply(strsplit(intergenic.mapping$locus, '-'), '[', 2))
ros.network.intergenic <- intergenic.mapping %>%
  filter(!is.na(upstream.gene)) %>%
  filter(upstream.gene %in% unique(full.network.with.evidence$Source) | upstream.gene %in% unique(full.network.with.evidence$Target))

ros.network.bayes.intergenic <- left_join(ros.network.intergenic, bayes) %>% filter(adj.p.value <= 0.05)
```

```{r}
# generating supplementary table S2
ros.network.bayes.genes <- ros.network.bayes.genes %>% dplyr::select(locus, bayes, adj.p.value)
ros.network.bayes.intergenic <- ros.network.bayes.intergenic %>% dplyr::select(locus, bayes, adj.p.value)

table.s2 <- rbind(ros.network.bayes.genes, ros.network.bayes.intergenic)
write_csv(table.s2, file = '../supplemental-tables/S2-OSR-network-Sig-BP-loci.csv')
```


Also assessing if there is an enrichment of significant bayes probability genes amongst OSR network

```{r}
# assessing if the final network is enriched for genes with significant bayes probability
# getting list of all loci in network to get population size and population successes
bayes.genic <- bayes[!grepl('-', bayes$locus), ]
all.genic.loci <- unique(bayes.genic$locus)
n.total <- length(all.genic.loci)
all.network.genes <- c(full.network.with.evidence$Source, full.network.with.evidence$Target)
all.network.genes <- all.network.genes[!is.na(all.network.genes)]
n.network.total <- length(unique(all.network.genes))
# getting sample size and sample successes
n.significant.loci <- length(unique(bayes.genic %>% filter(adj.p.value <= 0.05) %>% pull(locus)))
n.significant.network.loci <- nrow(ros.network.bayes.genes) + nrow(ros.network.bayes.intergenic)

# performing hypergeometric test to calculate significance of enrichment of intergenic loci amonst significant hits
p.value <- phyper(q = (n.significant.network.loci - 1),
                  m = n.network.total,
                  n = ( n.total - n.network.total ),
                  k = n.significant.loci,
                  lower.tail = FALSE)
print(paste('hypergeometric test p value for enrichment of significant bayes loci amongst OSR network nodes = ', p.value, sep = ''))

# getting additional values for printing
regulated.tfs <- intersect(x = full.network.with.evidence$Source, y = full.network.with.evidence$Target)
regulated.tfs <- regulated.tfs[!is.na(regulated.tfs)]

n.genes <- length(unique(full.network.with.evidence %>% filter(!Target %in% regulated.tfs, !is.na(Target)) %>% pull(Target)))
n.tfs <- length(unique(full.network.with.evidence %>% filter(!is.na(Source)) %>% pull(Source)))
n.sig.genes <- nrow(ros.network.bayes.genes)
n.sig.intergenics <- nrow(ros.network.bayes.intergenic)

print(paste('the final network contained ', n.genes, ' genes and ', n.tfs, ' TFs', sep = ''))
print(paste('the network was enriched for ', n.sig.genes, ' genes and ', n.sig.intergenics, ' intergenic regions with significant BP', sep = ''))
```

Next, looking to see if the nodes of the OSR network are enriched amongst genes that responded (enriched or depleted) to drug treatment in the CRISPRi dataset

This is very similar to the analysis done previously, looking for enrichment of significant BP genes within the CRISPRi datasets

```{r}
# performing hypergeometric tests to look for over-representation of OSR network genes enriched/depleted in CRISPRi data
# first getting gene names in OSR network, either by all nodes, all nodes + any missing of the 68 ROS genes, or only the 68 ROS genes
# all.osr.nodes <- unique(c(full.network.with.evidence$Source.rv.number, full.network.with.evidence$Target.rv.number))
all.osr.nodes <- unique(c(full.network.with.evidence$Source.rv.number, full.network.with.evidence$Target.rv.number, oxidative.stress.genes$rv.number))
# all.osr.nodes <- unique(oxidative.stress.genes$rv.number)

all.osr.nodes <- all.osr.nodes[!is.na(all.osr.nodes)]

# importing CRISPRi gene enrichment/depletion data from Supplemental Data 2
treatments <- c('BDQ 1d', 'BDQ 5d', 'BDQ 10d', 'CLR 1d', 'CLR 5d', 'CLR 10d', 'EMB 1d', 'EMB 5d', 'EMB 10d', 'INH 1d', 'INH 5d', 'INH 10d', 'LVX 1d', 'LVX 5d', 'LVX 10d',
                'LZD 1d', 'LZD 5d', 'LZD 10d', 'RIF 1d', 'RIF 5d', 'RIF 10d', 'STR 1d', 'STR 5d', 'VAN 1d', 'VAN 5d', 'VAN 10d')
sheets <- c(3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17,
            18, 19, 20, 21, 22, 23, 24, 25, 27, 28, 29)

crispri.osr.enrichment.test.list <- list()

for (d in 1:length(treatments)) {
  treatment <- treatments[d]
  selected.sheet <- sheets[d]
  
  crispri.data <- readxl::read_excel(path = '../data/Fig6/published-data/li-et-al-2022-table-SD-F1.xlsx', sheet = selected.sheet)
  # print(paste('treatment = ', treatment, ', sheet = ', selected.sheet, ', columns = ', ncol(crispri.data), sep = ''))
  
  crispri.data$`locus tag` <- paste('Rv', unlist(lapply(strsplit(unlist(lapply(strsplit(crispri.data$ORF_ID, ':'), '[', 1)), 'RVBD'), '[', 2)), sep = '')
  crispri.data[crispri.data == 'RvNA'] <- NA
  crispri.data$`gene name` <- unlist(lapply(strsplit(crispri.data$ORF_ID, ':'), '[', 2))
  crispri.data <- crispri.data %>% filter(`locus tag` %in% mtb.gene.map$locus)
  # high dose columns
  crispri.data.high <- crispri.data[, c(1, 22, 23, 26, 27, 28, 29)]
  names(crispri.data.high) <- c('ORF', 'neg l2fc', 'neg fdr', 'pos l2fc', 'pos fdr', 'locus tag', 'gene name')
  # med dose columns
  crispri.data.med <- crispri.data[, c(1, 14, 15, 18, 19, 28, 29)]
  names(crispri.data.med) <- c('ORF', 'neg l2fc', 'neg fdr', 'pos l2fc', 'pos fdr', 'locus tag', 'gene name')
  # low dose columns
  crispri.data.low <- crispri.data[, c(1, 6, 7, 10, 11, 28, 29)]
  names(crispri.data.low) <- c('ORF', 'neg l2fc', 'neg fdr', 'pos l2fc', 'pos fdr', 'locus tag', 'gene name')

  experiment.groups <- list(crispri.data.high, crispri.data.med, crispri.data.low)
  experiment.conds <- c('high', 'medium', 'low')
  crispri.hyper.tests.per.dose <- list()
  for (e in 1:length(experiment.groups)) {
    crispri.data.select <- experiment.groups[[e]]
    dose <- experiment.conds[e]
    # filtering enrichment/depletion data genes to only include genes that exist in established genome
    crispri.data.filtered <- crispri.data.select %>% filter(`locus tag` %in% unique(mtb.gene.map$locus))
    # getting population size for downstream hypergeometric tests
    population.size <- nrow(crispri.data.filtered)
    # getting input values for hypergeometric test
    sample.size <- length(all.osr.nodes)
    #################################################################################
    # hypergeometric test for all significant genes from crispri
    # extracting list of significant genes (crispri)
    crispri.significant.genes <- unique(rbind(crispri.data.filtered %>%
                                                filter(`pos l2fc` >= 1, `pos fdr` <= 0.01),
                                              crispri.data.filtered %>%
                                                filter(`neg l2fc` <= -1, `neg fdr` <= 0.01)) %>% pull(`locus tag`))
    crispri.significant.genes <- crispri.significant.genes[!is.na(crispri.significant.genes)]
    population.successes <- length(crispri.significant.genes)
    sample.successes <- length(intersect(x = all.osr.nodes, y = crispri.significant.genes))
    # saving results of hypergeometric test
    hyper.all <- tibble(
      dataset = 'CRISPRi',
      response = 'Responder',
      hyper.test.p.value = phyper(q = (sample.successes - 1),
                                  m = population.successes,
                                  n = ( population.size - population.successes ),
                                  k = sample.size,
                                  lower.tail = FALSE),
      population.size = population.size,
      n.osr.genes = sample.size,
      n.significant.exp.genes = population.successes, n.overlap = sample.successes,
      overlapping.genes = list(intersect(x = all.osr.nodes, y = crispri.significant.genes)))
    #################################################################################
    # hypergeometric test only for genes with alleviating or enriched interactions
    crispri.up.genes <- unique(crispri.data.filtered %>%
                                 filter(`pos l2fc` >= 1, `pos fdr` <= 0.01) %>% pull(`locus tag`))
    crispri.up.genes <- crispri.up.genes[!is.na(crispri.up.genes)]
    population.successes <- length(crispri.up.genes)
    sample.successes <- length(intersect(x = all.osr.nodes, y = crispri.up.genes))
    hyper.up <- tibble(
      dataset = 'CRISPRi',
      response = 'Enriched',
      hyper.test.p.value = phyper(q = (sample.successes - 1),
                                  m = population.successes,
                                  n = ( population.size - population.successes ),
                                  k = sample.size,
                                  lower.tail = FALSE),
      population.size = population.size,
      n.osr.genes = sample.size,
      n.significant.exp.genes = population.successes, n.overlap = sample.successes,
      overlapping.genes = list(intersect(x = all.osr.nodes, y = crispri.up.genes)))
    #################################################################################
    # hypergeometric test only for genes with aggravating or depleted interactions
    crispri.down.genes <- unique(crispri.data.filtered %>%
                                   filter(`neg l2fc` <= -1, `neg fdr` <= 0.01) %>% pull(`locus tag`))
    crispri.down.genes <- crispri.down.genes[!is.na(crispri.down.genes)]
    population.successes <- length(crispri.down.genes)
    sample.successes <- length(intersect(x = all.osr.nodes, y = crispri.down.genes))
    hyper.down <- tibble(
      dataset = 'CRISPRi',
      response = 'Depleted',
      hyper.test.p.value = phyper(q = (sample.successes - 1),
                                  m = population.successes,
                                  n = ( population.size - population.successes ),
                                  k = sample.size,
                                  lower.tail = FALSE),
      population.size = population.size,
      n.osr.genes = sample.size,
      n.significant.exp.genes = population.successes, n.overlap = sample.successes,
      overlapping.genes = list(intersect(x = all.osr.nodes, y = crispri.down.genes)))

    # aggregating all data from three hypergeometric tests
    hyper.tests <- rbind(hyper.all, hyper.up, hyper.down)
    hyper.tests$treatment <- paste(dose, ' dose ', treatment, sep = '')
    crispri.hyper.tests.per.dose[[e]] <- hyper.tests
  }
  # putting together all hypergeometric tests for CRISPRi
  crispri.hyper.tests <- do.call(rbind, crispri.hyper.tests.per.dose)
  crispri.osr.enrichment.test.list[[d]] <- crispri.hyper.tests
}

crispri.osr.enrichment.tests <- do.call(rbind, crispri.osr.enrichment.test.list)
# crispri.enrichment.tests$adj.p.value <- p.adjust(crispri.enrichment.tests$hyper.test.p.value, method = 'fdr')
crispri.osr.enrichment.tests.filtered <- crispri.osr.enrichment.tests %>% filter(hyper.test.p.value <= 0.05)
# crispri.osr.enrichment.tests.filtered$overlapping.genes <- sapply(crispri.osr.enrichment.tests.filtered$overlapping.genes, function(x) paste(x, collapse = ', '))
```

```{r}
write_csv(crispri.osr.enrichment.tests.filtered, file = '../data/Fig6/osr-network-enrichment-in-crispri.csv')
```


```{r}
# generating Table 2, using information from the significant BP genes enriched in CRISPRi and the OSR network enriched in CRISPRi
bayes.v.crispri <- crispri.enrichment.tests.filtered
osr.v.crispri <- crispri.osr.enrichment.tests.filtered
# summarize BP vs. CRISPRi
bayes.v.crispri.summary.list <- list()
for (r in 1:nrow(bayes.v.crispri)) {
  select.test <- bayes.v.crispri[r, ]
  gene.list.conversion <- tibble(rv.number = unlist(select.test$overlapping.genes))
  gene.list.conversion <- suppressMessages(left_join(gene.list.conversion, gene.name.conversion.table))
  bayes.v.crispri.summary.list[[r]] <- tibble(`BP vs. CRISPRi hypergeometric test p value` = select.test$hyper.test.p.value,
                                              `Sig. BP genes` = list(gene.list.conversion$locus),
                                              `CRISPRi treatment` = paste(select.test$response, ' in ', select.test$treatment, sep = ''))
}
bayes.v.crispri.summary <- do.call(rbind, bayes.v.crispri.summary.list)

# summarize OSR vs. CRISPRi
osr.v.crispri.summary.list <- list()
for (r in 1:nrow(osr.v.crispri)) {
  select.test <- osr.v.crispri[r, ]
  gene.list.conversion <- tibble(rv.number = unlist(select.test$overlapping.genes))
  gene.list.conversion <- suppressMessages(left_join(gene.list.conversion, gene.name.conversion.table))
  osr.v.crispri.summary.list[[r]] <- tibble(`OSR vs. CRISPRi hypergeometric test p value` = select.test$hyper.test.p.value,
                                            `OSR genes` = list(gene.list.conversion$locus),
                                            `CRISPRi treatment` = paste(select.test$response, ' in ', select.test$treatment, sep = ''))
}
osr.v.crispri.summary <- do.call(rbind, osr.v.crispri.summary.list)

# collecting treatment conditions in common between the two sets of enrichment tests
common.treatments <- intersect(x = bayes.v.crispri.summary$`CRISPRi treatment`, y = osr.v.crispri.summary$`CRISPRi treatment`)
common.test.tables <- list()
for (c in 1:length(common.treatments)) {
  common.bayes.test <- bayes.v.crispri.summary %>% filter(`CRISPRi treatment` == common.treatments[c])
  common.osr.test <- osr.v.crispri.summary %>% filter(`CRISPRi treatment` == common.treatments[c])
  three.way.genes <- intersect(x = unlist(common.bayes.test$`Sig. BP genes`),
                               y = unlist(common.osr.test$`OSR genes`))
  
  common.test.tables[[c]] <- tibble(`CRISPRi treatment` = common.treatments[c],
                                    `BP vs. CRISPRi hypergeometric test p value` = common.bayes.test$`BP vs. CRISPRi hypergeometric test p value`,
                                    `Sig. BP genes` = common.bayes.test$`Sig. BP genes`,
                                    `OSR vs. CRISPRi hypergeometric test p value` = common.osr.test$`OSR vs. CRISPRi hypergeometric test p value`,
                                    `OSR genes` = common.osr.test$`OSR genes`,
                                    `Three way overlapping genes` = list(three.way.genes))
}
common.tests <- do.call(rbind, common.test.tables)

# collecting significant tests unique to either the BP vs. CRISPRi or OSR vs. CRISPRi results
bayes.v.crispri.unique <- bayes.v.crispri.summary %>% filter(!`CRISPRi treatment` %in% common.treatments)
osr.v.crispri.unique <- osr.v.crispri.summary %>% filter(!`CRISPRi treatment` %in% common.treatments)

all.tests.summary <- bind_rows(common.tests, bayes.v.crispri.unique, osr.v.crispri.unique)

# splitting columns with lists of genes for exporting
all.tests.summary$`Sig. BP genes` <- sapply(all.tests.summary$`Sig. BP genes`, function(x) paste(x, collapse = ', '))
all.tests.summary$`OSR genes` <- sapply(all.tests.summary$`OSR genes`, function(x) paste(x, collapse = ', '))
all.tests.summary$`Three way overlapping genes` <- sapply(all.tests.summary$`Three way overlapping genes`, function(x) paste(x, collapse = ', '))

# further annotating table with its own information
all.tests.summary$`Gene Group` <- unlist(lapply(strsplit(all.tests.summary$`CRISPRi treatment`, ' in'), '[', 1))
all.tests.summary$Dose <- unlist(lapply(strsplit(unlist(lapply(strsplit(all.tests.summary$`CRISPRi treatment`, 'in '), '[', 2)), ' dose'), '[', 1))
all.tests.summary$Antibiotic <- unlist(lapply(strsplit(unlist(lapply(strsplit(all.tests.summary$`CRISPRi treatment`, 'dose '), '[', 2)), ' '), '[', 1))
all.tests.summary$`Day(s)` <- as.numeric(unlist(lapply(strsplit(unlist(lapply(strsplit(all.tests.summary$`CRISPRi treatment`, ' '), '[', 6)), 'd'), '[', 1)))

# factoring and reordering
all.tests.summary$Antibiotic <- factor(all.tests.summary$Antibiotic, levels = c('INH', 'EMB', 'CLR', 'RIF', 'STR', 'BDQ', 'LZD', 'VAN'))
all.tests.summary$Dose <- factor(all.tests.summary$Dose, levels = c('low', 'medium', 'high'))
all.tests.summary$`Gene Group` <- factor(all.tests.summary$`Gene Group`, levels = c('Enriched', 'Depleted', 'Responder'))

all.tests.summary <- all.tests.summary[with(all.tests.summary, order(Antibiotic, Dose, `Day(s)`, `Gene Group`)), ]
all.tests.summary <- all.tests.summary %>% dplyr::select(Antibiotic, Dose, `Day(s)`, `Gene Group`,
                                                         `BP vs. CRISPRi hypergeometric test p value`, `Sig. BP genes`,
                                                         `OSR vs. CRISPRi hypergeometric test p value`, `OSR genes`, `Three way overlapping genes`)
```

```{r}
write_csv(all.tests.summary, file = '../data/Fig6/Table2-crispri-enrichment-summary.csv')
```


Next, analyzing data from EGRIN - Environment and Gene Regulatory Influence Network - to compliment existing network information from TFOE and ChIP-seq

EGRIN enables the identification of regulatory relationships that may or may not be differentially active across different environments, including oxidative stress

EGRIN data was pulled from the following GitHub: https://github.com/baliga-lab/egrin2api_mtu/tree/master

```{r}
# reading in EGRIN data
# corem.targets <- read_csv(file = '../data/Fig6/published-data/egrin2api_mtu/data/corem_genes_20171129.csv')
# names(corem.targets) <- c('corem', 'gene')
# corem.targets$corem <- as.numeric(unlist(lapply(strsplit(corem.targets$corem, '_'), '[', 2)))
# # also evaluating which GREs are associated with each corem
# corem.gre <- read_csv(file = '../data/Fig6/published-data/egrin2api_mtu/data/20161223.GRE.corem.associations.csv')
# # also checking which condition blocks are associated with each corem
# corem.conditions <- read_csv(file = '../data/Fig6/published-data/egrin2api_mtu/data/corem_unique_condblocks-20170216.csv')
# names(corem.conditions)[1] <- 'corem'
# # first checking to see what corems have predicted regulons enriched for ROS genes
# # this is very similar to the analysis above, that used TFOE data
# # setting pop. size parameter for hypergeometric test
# population.size <- length(unique(corem.targets$gene))
# included.ros.genes <- oxidative.stress.genes %>% filter(rv.number %in% unique(corem.targets$gene))
# # performing hypergeometric tests to identify which corems are enriched for ROS genes
# unique.corems <- unique(corem.targets$corem)
# enrichment.data <- list()
# for (t in 1:length(unique.corems)) {
#   # filter gene lists for each corem to get regulon
#   corem.data <- corem.targets %>% filter(corem == unique.corems[t])
#   # setting remainder parameters for hypergeometric test
#   corem.genes <- unique(corem.data$gene)
#   select.corem.gres <- unique(corem.gre %>% filter(corem == unique.corems[t]) %>% pull(gre))
#   select.corem.gre.pvalues <- corem.gre %>% filter(corem == unique.corems[t]) %>% pull(qval_BH)
#   select.corem.conditions <- corem.conditions %>% filter(corem == unique.corems[t]) %>% pull(EGRIN2.block)
#   sample.size <- length(corem.genes)
#   population.successes <- nrow(included.ros.genes)
#   ros.genes <- included.ros.genes$rv.number
#   overlap <- intersect(x = corem.genes, y = ros.genes)
#   sample.successes <- length(overlap)
#   # saving results of hypergeometric test
#   hyper.test <- tibble(
#     corem = unique.corems[t],
#     n.corem.genes = sample.size,
#     # corem.genes = paste(corem.genes, collapse = ', '),
#     # corem.gres = paste(select.corem.gres, collapse = ', '),
#     # corem.gre.pvalues = paste(select.corem.gre.pvalues, collapse = ', '),
#     # corem.condition.blocks = paste(select.corem.conditions, collapse = ', '),
#     corem.genes = list(corem.genes),
#     corem.gres = list(select.corem.gres),
#     corem.gre.pvalues = list(select.corem.gre.pvalues),
#     corem.condition.blocks = list(select.corem.conditions),
#     n.ros.genes = population.successes, n.overlap = sample.successes,
#     overlapping.loci = list(overlap),
#     hyper.test.pvalue = phyper(q = (sample.successes - 1),
#                                m = population.successes,
#                                n = ( population.size - population.successes ),
#                                k = sample.size,
#                                lower.tail = FALSE))
#   enrichment.data[[t]] <- hyper.test
# }
# egrin.network.enrichment <- do.call(rbind, enrichment.data)
# egrin.network.enrichment$adj.pvalue <- p.adjust(egrin.network.enrichment$hyper.test.pvalue, method = 'fdr')
# # filtering for significant hypergeometric tests to find regulons enriched for ROS genes
# egrin.network.enrichment.filtered <- egrin.network.enrichment %>%
#   filter(hyper.test.pvalue <= 0.05)
# egrin.network.enrichment.filtered.fdr <- egrin.network.enrichment %>%
#   filter(adj.pvalue <= 0.05)
# # write_csv(egrin.network.enrichment.filtered.fdr, file = '../data/Fig6/oxidative-stress-network/ros-enriched-corems.csv')
# ##################################################################################
# # also getting a network that retains all corems with at least one ROS gene
# all.ros.corems <- corem.targets %>% filter(gene %in% oxidative.stress.genes$rv.number)
# unique.ros.corem.genes <- unique(all.ros.corems$gene)
# 
# network.overlap <- intersect(x = unique.ros.corem.genes, y = unique(c(full.network.with.evidence$Source.rv.number, full.network.with.evidence$Target.rv.number)))
# 
# corem.inclusion.information <- tibble(rv.number = unique.ros.corem.genes,
#                                       inclusion = ifelse(unique.ros.corem.genes %in% network.overlap, yes = 'gene in network and corem', no = 'gene not in network'))
# corem.inclusion.information <- left_join(corem.inclusion.information, oxidative.stress.genes)
```

```{r}
# # next, identifying which TFs are associated with each corem
# # hypergeometric test to evaluate if corem x is enriched for genes controlled by TF y
# # TF information based on either union or intersection of TFOE and ChIP-seq data
# # only including TFs and targets that are in corems
# unique.corem.genes <- unique(corem.targets$gene)
# tfoe.chipseq.union <- tfoe.chipseq %>% filter(adj.pvalue <= 0.05 | !is.na(Evidence)) %>% filter(TF %in% unique.corem.genes | Gene %in% unique.corem.genes)
# tfoe.chipseq.intersection <- tfoe.chipseq %>% filter(adj.pvalue <= 0.05, !is.na(Evidence)) %>% filter(TF %in% unique.corem.genes | Gene %in% unique.corem.genes)
# 
# unique.tfs.union <- unique(tfoe.chipseq.union$TF)
# unique.tfs.intersection <- unique(tfoe.chipseq.intersection$TF)
# 
# population.size <- length(unique.corem.genes)
# 
# corem.enrichment.tables <- list()
# for (i in 1:nrow(egrin.network.enrichment.filtered.fdr)) {
#   select.corem.data <- egrin.network.enrichment.filtered.fdr[i, ]
#   select.corem.id <- select.corem.data$corem
#   select.corem.genes <- unlist(select.corem.data$corem.genes)
#   sample.size <- length(select.corem.genes)
#   
#   # enrichment test for union of tfoe or chip-seq data
#   union.enrichment.tables <- list()
#   for (t in 1:length(unique.tfs.union)) {
#     select.tfoe.chipseq.union <- tfoe.chipseq.union %>% filter(TF == unique.tfs.union[t])
#     union.population.successes <- nrow(select.tfoe.chipseq.union)
#     tfoe.chipseq.genes <- unique(select.tfoe.chipseq.union$Gene)
#     overlap <- intersect(x = select.corem.genes, y = tfoe.chipseq.genes)
#     sample.successes <- length(overlap)
#     
#     # saving results of hypergeometric test
#     suppressWarnings(hyper.test <- tibble(
#       corem = select.corem.id,
#       n.corem.genes = sample.size,
#       corem.genes = select.corem.data$corem.genes,
#       corem.gres = select.corem.data$corem.gres,
#       corem.gre.pvalues = select.corem.data$corem.gre.pvalues,
#       TF = unique.tfs.union[t],
#       n.tfoe.chipseq.genes = union.population.successes, n.overlap = sample.successes,
#       overlapping.loci = list(overlap),
#       hyper.test.pvalue = phyper(q = (sample.successes - 1),
#                                  m = union.population.successes,
#                                  n = ( population.size - union.population.successes ),
#                                  k = sample.size,
#                                  lower.tail = FALSE)))
#     union.enrichment.tables[[t]] <- hyper.test
#   }
#   union.enrichment <- do.call(rbind, union.enrichment.tables)
#   union.enrichment$adj.pvalue <- p.adjust(union.enrichment$hyper.test.pvalue, method = 'fdr')
#   union.enrichment$comparison <- 'union tfoe or chip-seq'
#   
#   # enrichment tests for intersection of tfoe and chip-seq
#   intersection.enrichment.tables <- list()
#   for (s in 1:length(unique.tfs.intersection)) {
#     select.tfoe.chipseq.intersection <- tfoe.chipseq.intersection %>% filter(TF == unique.tfs.intersection[s])
#     intersection.population.successes <- nrow(select.tfoe.chipseq.intersection)
#     tfoe.chipseq.genes <- unique(select.tfoe.chipseq.intersection$Gene)
#     overlap <- intersect(x = select.corem.genes, y = tfoe.chipseq.genes)
#     sample.successes <- length(overlap)
#     
#     # saving results of hypergeometric test
#     suppressWarnings(hyper.test <- tibble(
#       corem = select.corem.id,
#       n.corem.genes = sample.size,
#       corem.genes = select.corem.data$corem.genes,
#       corem.gres = select.corem.data$corem.gres,
#       corem.gre.pvalues = select.corem.data$corem.gre.pvalues,
#       TF = unique.tfs.intersection[s],
#       n.tfoe.chipseq.genes = intersection.population.successes, n.overlap = sample.successes,
#       overlapping.loci = list(overlap),
#       hyper.test.pvalue = phyper(q = (sample.successes - 1),
#                                  m = intersection.population.successes,
#                                  n = ( population.size - intersection.population.successes ),
#                                  k = sample.size,
#                                  lower.tail = FALSE)))
#     intersection.enrichment.tables[[s]] <- hyper.test
#   }
#   intersection.enrichment <- do.call(rbind, intersection.enrichment.tables)
#   intersection.enrichment$adj.pvalue <- p.adjust(intersection.enrichment$hyper.test.pvalue, method = 'fdr')
#   intersection.enrichment$comparison <- 'intersection tfoe and chip-seq'
#   
#   corem.enrichment.tables[[i]] <- rbind(union.enrichment, intersection.enrichment)
# }
# 
# corem.tf.enrichment <- do.call(rbind, corem.enrichment.tables)
# corem.tf.enrichment.filtered <- corem.tf.enrichment %>% filter(hyper.test.pvalue <= 0.05)
# corem.tf.enrichment.filtered.fdr <- corem.tf.enrichment %>% filter(adj.pvalue <= 0.05)
```

```{r}
# # swapping Rv number with gene name to compare TFs associated with each corem to those in the existing network
# corem.tfs <- tibble(corem = corem.tf.enrichment.filtered.fdr$corem,
#                     rv.number = corem.tf.enrichment.filtered.fdr$TF)
# corem.tfs <- left_join(corem.tfs, gene.name.conversion.table)
# unique.corem.tfs <- unique(corem.tfs$locus)
# 
# overlapping.tfs <- intersect(x = unique.corem.tfs, y = unique(full.network.with.evidence$Source))
```

```{r}
# next, looking at remainder files from EGRIN2 analysis, including TSS information for each gene
# tss.info <- read_csv(file = '../data/Fig6/published-data/egrin2api_mtu/data/20161006.gene.TSS.start.stop.csv')
# # reading in TF binding position list, as determined by ChIP-seq
# tf.binding.positions.chip.seq <- read_csv(file = '../data/Fig6/published-data/egrin2api_mtu/data/chip_seq_data.csv')
# tf.binding.positions.chip.seq.filtered <- tf.binding.positions.chip.seq %>% filter(tf %in% unique(corem.tf.enrichment.filtered.fdr$TF))
# 
# # also assessing which conditions are associated with each corem enriched for ROS genes
# corem.conditions.filtered <- corem.conditions %>% filter(corem %in% unique(corem.tf.enrichment.filtered.fdr$corem))
# 
# # reading in Mtb H37Rv reference genome from: https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_000195955.2/
# # h37rv.genome <- read.fasta(file = '../data/genome-annotations/GCF_000195955.2_ASM19595v2_genomic.fna', seqtype = 'DNA')
# h37rv.genome = readDNAStringSet('../data/genome-annotations/GCF_000195955.2_ASM19595v2_genomic.fna')
```

```{r}
# # writing code to extract the -200 to +100 region of each gene's TSS
# # will be looking specifically at the regulatory region of genes who are controlled by TFs identified above
# 
# # first getting Rv numbers for each TF
# unique.tfs <- tibble(locus = unique.corem.tfs)
# unique.tfs <- left_join(unique.tfs, gene.name.conversion.table)
# unique.corem.tfs.rv <- unique.tfs$rv.number
# 
# for (t in 1:length(unique.corem.tfs.rv)) {
#   tf <- unique.corem.tfs.rv[t]
#   tf.regulon <- tfoe.chipseq.union %>% filter(TF == tf)
#   regulon.genes <- unique(tf.regulon$Gene)
#   
#   regulon.gene.seqs <- list()
#   for (r in 1:length(regulon.genes)) {
#     gene <- regulon.genes[r]
#     gene.tss.info <- tss.info %>% filter(gene == regulon.genes[r])
#     gene.tss.position <- gene.tss.info %>% pull(genome.position)
#     if (gene.tss.info$strand == '+') {
#       regulated.region <- subseq(h37rv.genome,
#                                  start = gene.tss.position - 200,
#                                  end = gene.tss.position + 100)
#     }
#     else {
#       regulated.region <- subseq(h37rv.genome,
#                                  start = gene.tss.position - 100,
#                                  end = gene.tss.position + 200)
#     }
#     names(regulated.region) <- gene
#     regulon.gene.seqs[[r]] <- regulated.region
#   }
#   # combine all subseqs into one DNAStringSet
#   seqs <- do.call(c, regulon.gene.seqs)
#   # write multi-fasta
#   writeXStringSet(seqs,
#                   filepath = paste0('../data/Fig6/oxidative-stress-network/ros-corem-tf-regulon-mfastas/TF-', tf, '-regulon.fasta'),
#                   format = "fasta")
# }
# # the files generated here can be input into MEME to identify motif consensus sequences within each regulon for comparison to GREs
```






