---
title: "Figure 7 Notebook: Generation of Oxidative Stress Response (OSR) Network"
output: html_notebook
---

**Author: Evan Pepper-Tunick**

Last updated: Mar 27, 2026

This notebook is for developing an oxidative stress transcriptional regulatory network in M. tuberculosis. To do this, two datasets will primarily be used:

1. RNA-seq data from genome wide transcription factor over expression (TFOE) in Mtb: Rustad, T. R. et al. Mapping and manipulating the Mycobacterium tuberculosis transcriptome using a transcription factor overexpression-derived regulatory network.
2. ChIP-seq data from genome wide TFOE in Mtb: Minch, K. J. et al. The DNA-binding network of Mycobacterium tuberculosis & Turkarslan, S. et al. A comprehensive map of genome-wide gene regulation in Mycobacterium tuberculosis.

**Important note: to run notebook, user's computer must have ≥ 4.0 GB memory**

```{r}
# importing necessary libraries
suppressMessages(
  {
  library(ggplot2)
  library(ggridges)
  library(ggupset)
  library(tidyverse)
  library(purrr)
  library(drc)
  library(knitr)
  library(factoextra)
  library(naniar)
  library(rstatix)
  library(ggpubr)
  library(scales)
  library(ggrepel)
  library(see)
  library(viridis)
  library(seqinr)
  library(Biostrings)
  }
)
```

```{r}
# reading in gene name mapping files
mtb.gene.map <- read_csv(file = '../data/genome-annotations/mtb-h37rv-full-genome-annotation.csv', show_col_types = FALSE)
gene.name.conversion.table <- read_csv(file = '../data/genome-annotations/gene-locus-map.csv')
```

# assembling lists of oxidative stress related genes using Gene Ontology categories, in order to develop OSR network:
antioxidant activity --> GO:0016209
cellular oxidant detoxification --> GO:0098869
removal of superoxide radicals --> GO:0019430
mycothiol-dependent detoxification --> GO:0010127
response to oxidative stress --> GO:0006979
response to nitrosative stress --> GO:0051409

```{r}
#################################################################################
go.1 <- readxl::read_excel(path = '../data/Fig7/oxidative-stress-network/oxidative-stress-genes.xlsx', sheet = 2)
go.2 <- readxl::read_excel(path = '../data/Fig7/oxidative-stress-network/oxidative-stress-genes.xlsx', sheet = 3)
go.3 <- readxl::read_excel(path = '../data/Fig7/oxidative-stress-network/oxidative-stress-genes.xlsx', sheet = 4)
go.4 <- readxl::read_excel(path = '../data/Fig7/oxidative-stress-network/oxidative-stress-genes.xlsx', sheet = 5)
go.5 <- readxl::read_excel(path = '../data/Fig7/oxidative-stress-network/oxidative-stress-genes.xlsx', sheet = 6)
go.6 <- readxl::read_excel(path = '../data/Fig7/oxidative-stress-network/oxidative-stress-genes.xlsx', sheet = 7)
go.data <- rbind(go.1, go.2, go.3, go.4, go.5, go.6)
go.data.filtered <- go.data %>% filter(`Gene Names (ordered locus)` %in% mtb.gene.map$locus)
go.data.filtered <- go.data.filtered %>% dplyr::select(`Gene Names (ordered locus)`, `GO Term`)
names(go.data.filtered) <- c('rv.number', 'GO Term')
oxidative.stress.genes <- go.data.filtered[match(unique(go.data.filtered$rv.number), go.data.filtered$rv.number), ]
oxidative.stress.genes <- left_join(oxidative.stress.genes, gene.name.conversion.table)
```

```{r}
# looking at TFOE data to extract regulons for oxidative stress network
tfoe.fc <- read_csv(file = '../data/Fig7/oxidative-stress-network/tfoe-fold-change.csv')
tfoe.fc <- tfoe.fc %>% pivot_longer(cols = -1, names_to = 'TF', values_to = 'FC')
tfoe.pval <- read_csv(file = '../data/Fig7/oxidative-stress-network/tfoe-p-values.csv')
tfoe.pval <- tfoe.pval %>% pivot_longer(cols = -1, names_to = 'TF', values_to = 'pvalue')
# column binding data
tfoe.data <- left_join(x = tfoe.fc, tfoe.pval)
tfoe.data$interaction <- ifelse(tfoe.data$FC < 0, 'repression', 'activation')
# reading in ChIP seq data for TF binding evidence
chip.seq <- read_csv(file = '../data/Fig7/oxidative-stress-network/chipseq-binding-evidence.csv')
chip.seq <- chip.seq %>% pivot_longer(cols = -1, names_to = 'TF', values_to = 'Evidence')
chip.seq.evidence <- na.omit(chip.seq)
# merging all tfoe and chipseq data
tfoe.chipseq <- left_join(tfoe.data, chip.seq)
tfoe.chipseq$adj.pvalue <- p.adjust(tfoe.chipseq$pvalue, method = 'fdr')
tfoe.chipseq$log10adj.pvalue <- log10(tfoe.chipseq$adj.pvalue)
# adding minimum pvalue to plot in log scale
tfoe.data$pvalue <- tfoe.data$pvalue + 7.46e-275
tfoe.data$adj.pvalue <- p.adjust(tfoe.data$pvalue, method = 'fdr')
tfoe.data$log10adj.pvalue <- log10(tfoe.data$adj.pvalue)

# printing out values for manuscript
print(paste('The RNAseq TFOE dataset contains information on ', length(unique(tfoe.data$TF)), ' TFs', sep = ''))
print(paste('The ChIP-seq TFOE dataset contains information on ', length(unique(chip.seq$TF)), ' TFs', sep = ''))
```

```{r}
# filtering data to only include significant interactions
tfoe.data.filtered <- tfoe.data %>% filter(adj.pvalue <= 0.05)
# setting pop. size parameter for hypergeometric test
population.size <- length(unique(tfoe.data.filtered$Gene))
# performing hypergeometric tests to identify which regulons are enriched for ROS genes
unique.tfs <- unique(tfoe.data.filtered$TF)
enrichment.data <- list()
for (t in 1:length(unique.tfs)) {
  # filter gene lists for each TF to get regulon
  regulon.data <- tfoe.data.filtered %>% filter(TF == unique.tfs[t])
  # setting remainder parameters for hypergeometric test
  regulated.genes <- unique(regulon.data$Gene)
  sample.size <- length(regulated.genes)
  population.successes <- nrow(oxidative.stress.genes)
  ros.genes <- oxidative.stress.genes$rv.number
  sample.successes <- length(intersect(x = regulated.genes, y = ros.genes))
  # saving results of hypergeometric test
  hyper.test <- tibble(
    TF = unique.tfs[t],
    n.regulated.genes = sample.size,
    n.ros.genes = population.successes, n.overlap = sample.successes,
    overlapping.loci = list(intersect(x = regulated.genes, y = ros.genes)),
    hyper.test.pvalue = phyper(q = (sample.successes - 1),
                               m = population.successes,
                               n = ( population.size - population.successes ),
                               k = sample.size,
                               lower.tail = FALSE))
  enrichment.data[[t]] <- hyper.test
}
network.enrichment <- do.call(rbind, enrichment.data)
network.enrichment$adj.pvalue <- p.adjust(network.enrichment$hyper.test.pvalue, method = 'fdr')
# filtering for significant hypergeometric tests to find regulons enriched for ROS genes
network.enrichment.filtered <- network.enrichment %>%
  filter(hyper.test.pvalue <= 0.05)
# capturing list of TFs with ROS-enriched regulons
ros.regulon.tfs <- unique(network.enrichment.filtered$TF)
# filter TFOE interactions for TFs identified from hypergeometric tests
tfoe.data.filtered.enriched <- tfoe.data.filtered %>% filter(TF %in% unique(network.enrichment.filtered$TF))
# prune network to only include ros genes
ros.network <- tfoe.data.filtered.enriched %>% filter(Gene %in% oxidative.stress.genes$rv.number)
# identifying the TFs with influence over the primary TFs of ROS genes
tfs.of.tfs <- tfoe.data.filtered %>% filter(Gene %in% unique(tfoe.data.filtered.enriched$TF))
# adding TFs of the TFs in the network back to the network itself
full.network <- rbind(ros.network, tfs.of.tfs) %>% arrange(TF)
# merging with ChIP-seq data to include interactions with evidence of binding
full.network.with.evidence <- left_join(full.network, chip.seq.evidence)
# adding in relationships from ChIPseq and TFOE evidence that was missed by enrichment analysis
# including all ROS gene interactions with evidence in both datasets
additional.data <- tfoe.chipseq %>% filter(Gene %in% oxidative.stress.genes$rv.number | TF %in% oxidative.stress.genes$rv.number,
                                                adj.pvalue <= 0.05, !is.na(Evidence))

# additional.data <- tfoe.chipseq %>% filter(Gene %in% oxidative.stress.genes$rv.number | TF %in% oxidative.stress.genes$rv.number,
#                                            !is.na(Evidence))

# removing instances of autoregulation
full.network.with.evidence <- full.network.with.evidence %>% filter(Gene != TF)
# merging with identified data
full.network.with.evidence <- rbind(full.network.with.evidence, additional.data)
full.network.with.evidence <- unique(full.network.with.evidence)
```

```{r}
# adding back into the network lone nodes of genes from the GO analysis that did not end up in the OSR network above
included.osr.genes <- intersect(x = c(full.network.with.evidence$Gene, full.network.with.evidence$TF), y = oxidative.stress.genes$rv.number)
extra.osr.nodes <- oxidative.stress.genes %>% filter(!rv.number %in% included.osr.genes)

# writing code to identify if specific osr gene is a Gene or TF based on inclusion in TFOE
tfoe.tfs <- unique(tfoe.data$TF)
extra.node.info.list <- list()
for (e in 1:nrow(extra.osr.nodes)) {
  selected.node <- extra.osr.nodes$rv.number[e]
  if (selected.node %in% tfoe.tfs) {
    extra.node.info.list[[e]] <- tibble(Gene = NA, TF = selected.node, FC = NA, pvalue = NA,
                                        interaction = NA, adj.pvalue = NA, log10adj.pvalue = NA, Evidence = NA)
  }
  else {
    extra.node.info.list[[e]] <- tibble(Gene = selected.node, TF = NA, FC = NA, pvalue = NA,
                                        interaction = NA, adj.pvalue = NA, log10adj.pvalue = NA, Evidence = NA)
  }
}
extra.node.info <- do.call(rbind, extra.node.info.list)
full.network.with.evidence <- rbind(full.network.with.evidence, extra.node.info)
```

```{r}
# annotating TFOE network
names(full.network.with.evidence)[1] <- 'rv.number'
full.network.with.evidence <- left_join(full.network.with.evidence, gene.name.conversion.table)
names(full.network.with.evidence)[c(1, 2, 9)] <- c('target.rv.number', 'rv.number', 'Target')
full.network.with.evidence <- left_join(full.network.with.evidence, gene.name.conversion.table)
names(full.network.with.evidence)[c(2, 10)] <- c('source.rv.number', 'Source')

full.network.with.evidence$Target <- ifelse(is.na(full.network.with.evidence$Target),
                                            full.network.with.evidence$target.rv.number,
                                            full.network.with.evidence$Target)

full.network.with.evidence$Source <- ifelse(is.na(full.network.with.evidence$Source),
                                            full.network.with.evidence$source.rv.number,
                                            full.network.with.evidence$Source)

# adding GO annotations in one column for either target, source, or both
annotations <- list()
for (r in 1:nrow(full.network.with.evidence)) {
  data <- full.network.with.evidence[r, ]
  if (data$target.rv.number %in% oxidative.stress.genes$rv.number & data$source.rv.number %in% oxidative.stress.genes$rv.number) {
    target.term <- oxidative.stress.genes %>% filter(rv.number == data$target.rv.number) %>% pull(`GO Term`)
    source.term <- oxidative.stress.genes %>% filter(rv.number == data$source.rv.number) %>% pull(`GO Term`)
    annotations[r] <- paste('Target: ', target.term, '; Source: ', source.term, sep = '')
  }
  else if (data$target.rv.number %in% oxidative.stress.genes$rv.number & !data$source.rv.number %in% oxidative.stress.genes$rv.number) {
    target.term <- oxidative.stress.genes %>% filter(rv.number == data$target.rv.number) %>% pull(`GO Term`)
    annotations[r] <- paste('Target: ', target.term, sep = '')
  }
  else if (!data$target.rv.number %in% oxidative.stress.genes$rv.number & data$source.rv.number %in% oxidative.stress.genes$rv.number) {
    source.term <- oxidative.stress.genes %>% filter(rv.number == data$source.rv.number) %>% pull(`GO Term`)
    annotations[r] <- paste('Source: ', source.term, sep = '')
  }
  else {
    annotations[r] <- 'No association'
  }
}

full.network.with.evidence$`GO annotation` <- unlist(annotations)

names(full.network.with.evidence) <- c('Target.rv.number', 'Source.rv.number', 'TFOE log2 FC', 'PValue', 'Interaction',
                                       'Adj.PValue', 'Adj.log10PValue', 'ChIP-seq Evidence', 'Target', 'Source',
                                       'GO annotation')
full.network.with.evidence <- full.network.with.evidence %>%
  dplyr::select(Source, Target, `TFOE log2 FC`, PValue, Adj.PValue, Interaction, `ChIP-seq Evidence`, `GO annotation`, Source.rv.number, Target.rv.number)
full.network.with.evidence$`Abs TFOE log2 FC` <- abs(full.network.with.evidence$`TFOE log2 FC`)

full.network.with.evidence <- unique(full.network.with.evidence)

# resplitting data between interaction network and lone OSR nodes
interaction.network <- full.network.with.evidence %>% filter(!is.na(Interaction))
lone.osr.nodes <- full.network.with.evidence %>% filter(is.na(Interaction))
```

```{r}
# writing to file
write_csv(interaction.network, file = '../data/Fig7/oxidative-stress-network/cytoscape-ros-network-input.csv')
write_csv(lone.osr.nodes, file = '../data/Fig7/oxidative-stress-network/cytoscape-ros-lone-nodes-input.csv')
```

Reading in calculated bayes probabilities with p values

```{r}
bayes <- read_csv(file = '../data/Fig6/LDF/bayes-probabilities/INH-Global-PValues.csv', show_col_types = FALSE)
```

```{r}
# analysis of genes from oxidative stress network and bayes
ros.network.bayes.genes <- bayes %>% filter(locus %in% full.network.with.evidence$Source | locus %in% full.network.with.evidence$Target,
                                            adj.p.value <= 0.05)
# looking separately at intergenic loci
intergenic.mapping <- gene.name.conversion.table
names(intergenic.mapping) <- c('full.locus', 'locus')
intergenic.mapping$upstream.gene <- unlist(lapply(strsplit(intergenic.mapping$locus, '-'), '[', 2))
ros.network.intergenic <- intergenic.mapping %>%
  filter(!is.na(upstream.gene)) %>%
  filter(upstream.gene %in% unique(full.network.with.evidence$Source) | upstream.gene %in% unique(full.network.with.evidence$Target))

ros.network.bayes.intergenic <- left_join(ros.network.intergenic, bayes) %>% filter(adj.p.value <= 0.05)
```

```{r}
# generating supplementary data 1
ros.network.bayes.genes <- ros.network.bayes.genes %>% dplyr::select(locus, bayes, adj.p.value)
ros.network.bayes.intergenic <- ros.network.bayes.intergenic %>% dplyr::select(locus, bayes, adj.p.value)

table.s1 <- rbind(ros.network.bayes.genes, ros.network.bayes.intergenic)
```

```{r}
write_csv(table.s1, file = '../supplemental-tables/Supplementary-Data-1-raw.csv')
```


Also assessing if there is an enrichment of significant bayes probability genes amongst OSR network

```{r}
# assessing if the final network is enriched for genes with significant bayes probability
# getting list of all loci in network to get population size and population successes
bayes.genic <- bayes[!grepl('-', bayes$locus), ]
all.genic.loci <- unique(bayes.genic$locus)
n.total <- length(all.genic.loci)
all.network.genes <- c(full.network.with.evidence$Source, full.network.with.evidence$Target)
all.network.genes <- all.network.genes[!is.na(all.network.genes)]
n.network.total <- length(unique(all.network.genes))
# getting sample size and sample successes
n.significant.loci <- length(unique(bayes.genic %>% filter(adj.p.value <= 0.05) %>% pull(locus)))
n.significant.network.loci <- nrow(ros.network.bayes.genes) + nrow(ros.network.bayes.intergenic)

# performing hypergeometric test to calculate significance of enrichment of intergenic loci amonst significant hits
p.value <- phyper(q = (n.significant.network.loci - 1),
                  m = n.network.total,
                  n = ( n.total - n.network.total ),
                  k = n.significant.loci,
                  lower.tail = FALSE)
print(paste('hypergeometric test p value for enrichment of significant bayes loci amongst OSR network nodes = ', p.value, sep = ''))

# getting additional values for printing
regulated.tfs <- intersect(x = full.network.with.evidence$Source, y = full.network.with.evidence$Target)
regulated.tfs <- regulated.tfs[!is.na(regulated.tfs)]

n.genes <- length(unique(full.network.with.evidence %>% filter(!Target %in% regulated.tfs, !is.na(Target)) %>% pull(Target)))
n.tfs <- length(unique(full.network.with.evidence %>% filter(!is.na(Source)) %>% pull(Source)))
n.sig.genes <- nrow(ros.network.bayes.genes)
n.sig.intergenics <- nrow(ros.network.bayes.intergenic)

print(paste('the final network contained ', n.genes, ' genes and ', n.tfs, ' TFs', sep = ''))
print(paste('the network was enriched for ', n.sig.genes, ' genes and ', n.sig.intergenics, ' intergenic regions with significant BP', sep = ''))
```

Next, looking to see if the nodes of the OSR network are enriched amongst genes that responded (enriched or depleted) to drug treatment in the CRISPRi dataset

This is very similar to the analysis done previously, looking for enrichment of significant BP genes within the CRISPRi datasets

```{r}
# performing hypergeometric tests to look for over-representation of OSR network genes enriched/depleted in CRISPRi data
# first getting gene names in OSR network, either by all nodes, all nodes + any missing of the 68 ROS genes, or only the 68 ROS genes
# all.osr.nodes <- unique(c(full.network.with.evidence$Source.rv.number, full.network.with.evidence$Target.rv.number))
all.osr.nodes <- unique(c(full.network.with.evidence$Source.rv.number, full.network.with.evidence$Target.rv.number, oxidative.stress.genes$rv.number))
# all.osr.nodes <- unique(oxidative.stress.genes$rv.number)

all.osr.nodes <- all.osr.nodes[!is.na(all.osr.nodes)]

# importing CRISPRi gene enrichment/depletion data from Supplemental Data 2
treatments <- c('BDQ 1d', 'BDQ 5d', 'BDQ 10d', 'CLR 1d', 'CLR 5d', 'CLR 10d', 'EMB 1d', 'EMB 5d', 'EMB 10d', 'INH 1d', 'INH 5d', 'INH 10d', 'LVX 1d', 'LVX 5d', 'LVX 10d',
                'LZD 1d', 'LZD 5d', 'LZD 10d', 'RIF 1d', 'RIF 5d', 'RIF 10d', 'STR 1d', 'STR 5d', 'VAN 1d', 'VAN 5d', 'VAN 10d')
sheets <- c(3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17,
            18, 19, 20, 21, 22, 23, 24, 25, 27, 28, 29)

crispri.osr.enrichment.test.list <- list()

for (d in 1:length(treatments)) {
  treatment <- treatments[d]
  selected.sheet <- sheets[d]
  
  crispri.data <- readxl::read_excel(path = '../data/published-data/li-et-al-2022-table-SD-F1.xlsx', sheet = selected.sheet)
  # print(paste('treatment = ', treatment, ', sheet = ', selected.sheet, ', columns = ', ncol(crispri.data), sep = ''))
  
  crispri.data$`locus tag` <- paste('Rv', unlist(lapply(strsplit(unlist(lapply(strsplit(crispri.data$ORF_ID, ':'), '[', 1)), 'RVBD'), '[', 2)), sep = '')
  crispri.data[crispri.data == 'RvNA'] <- NA
  crispri.data$`gene name` <- unlist(lapply(strsplit(crispri.data$ORF_ID, ':'), '[', 2))
  crispri.data <- crispri.data %>% filter(`locus tag` %in% mtb.gene.map$locus)
  # high dose columns
  crispri.data.high <- crispri.data[, c(1, 22, 23, 26, 27, 28, 29)]
  names(crispri.data.high) <- c('ORF', 'neg l2fc', 'neg fdr', 'pos l2fc', 'pos fdr', 'locus tag', 'gene name')
  # med dose columns
  crispri.data.med <- crispri.data[, c(1, 14, 15, 18, 19, 28, 29)]
  names(crispri.data.med) <- c('ORF', 'neg l2fc', 'neg fdr', 'pos l2fc', 'pos fdr', 'locus tag', 'gene name')
  # low dose columns
  crispri.data.low <- crispri.data[, c(1, 6, 7, 10, 11, 28, 29)]
  names(crispri.data.low) <- c('ORF', 'neg l2fc', 'neg fdr', 'pos l2fc', 'pos fdr', 'locus tag', 'gene name')

  experiment.groups <- list(crispri.data.high, crispri.data.med, crispri.data.low)
  experiment.conds <- c('high', 'medium', 'low')
  crispri.hyper.tests.per.dose <- list()
  for (e in 1:length(experiment.groups)) {
    crispri.data.select <- experiment.groups[[e]]
    dose <- experiment.conds[e]
    # filtering enrichment/depletion data genes to only include genes that exist in established genome
    crispri.data.filtered <- crispri.data.select %>% filter(`locus tag` %in% unique(mtb.gene.map$locus))
    # getting population size for downstream hypergeometric tests
    population.size <- nrow(crispri.data.filtered)
    # getting input values for hypergeometric test
    sample.size <- length(all.osr.nodes)
    #################################################################################
    # hypergeometric test for all significant genes from crispri
    # extracting list of significant genes (crispri)
    crispri.significant.genes <- unique(rbind(crispri.data.filtered %>%
                                                filter(`pos l2fc` >= 1, `pos fdr` <= 0.01),
                                              crispri.data.filtered %>%
                                                filter(`neg l2fc` <= -1, `neg fdr` <= 0.01)) %>% pull(`locus tag`))
    crispri.significant.genes <- crispri.significant.genes[!is.na(crispri.significant.genes)]
    population.successes <- length(crispri.significant.genes)
    sample.successes <- length(intersect(x = all.osr.nodes, y = crispri.significant.genes))
    # saving results of hypergeometric test
    hyper.all <- tibble(
      dataset = 'CRISPRi',
      response = 'Responder',
      hyper.test.p.value = phyper(q = (sample.successes - 1),
                                  m = population.successes,
                                  n = ( population.size - population.successes ),
                                  k = sample.size,
                                  lower.tail = FALSE),
      population.size = population.size,
      n.osr.genes = sample.size,
      n.significant.exp.genes = population.successes, n.overlap = sample.successes,
      overlapping.genes = list(intersect(x = all.osr.nodes, y = crispri.significant.genes)))
    #################################################################################
    # hypergeometric test only for genes with alleviating or enriched interactions
    crispri.up.genes <- unique(crispri.data.filtered %>%
                                 filter(`pos l2fc` >= 1, `pos fdr` <= 0.01) %>% pull(`locus tag`))
    crispri.up.genes <- crispri.up.genes[!is.na(crispri.up.genes)]
    population.successes <- length(crispri.up.genes)
    sample.successes <- length(intersect(x = all.osr.nodes, y = crispri.up.genes))
    hyper.up <- tibble(
      dataset = 'CRISPRi',
      response = 'Enriched',
      hyper.test.p.value = phyper(q = (sample.successes - 1),
                                  m = population.successes,
                                  n = ( population.size - population.successes ),
                                  k = sample.size,
                                  lower.tail = FALSE),
      population.size = population.size,
      n.osr.genes = sample.size,
      n.significant.exp.genes = population.successes, n.overlap = sample.successes,
      overlapping.genes = list(intersect(x = all.osr.nodes, y = crispri.up.genes)))
    #################################################################################
    # hypergeometric test only for genes with aggravating or depleted interactions
    crispri.down.genes <- unique(crispri.data.filtered %>%
                                   filter(`neg l2fc` <= -1, `neg fdr` <= 0.01) %>% pull(`locus tag`))
    crispri.down.genes <- crispri.down.genes[!is.na(crispri.down.genes)]
    population.successes <- length(crispri.down.genes)
    sample.successes <- length(intersect(x = all.osr.nodes, y = crispri.down.genes))
    hyper.down <- tibble(
      dataset = 'CRISPRi',
      response = 'Depleted',
      hyper.test.p.value = phyper(q = (sample.successes - 1),
                                  m = population.successes,
                                  n = ( population.size - population.successes ),
                                  k = sample.size,
                                  lower.tail = FALSE),
      population.size = population.size,
      n.osr.genes = sample.size,
      n.significant.exp.genes = population.successes, n.overlap = sample.successes,
      overlapping.genes = list(intersect(x = all.osr.nodes, y = crispri.down.genes)))

    # aggregating all data from three hypergeometric tests
    hyper.tests <- rbind(hyper.all, hyper.up, hyper.down)
    hyper.tests$treatment <- paste(dose, ' dose ', treatment, sep = '')
    crispri.hyper.tests.per.dose[[e]] <- hyper.tests
  }
  # putting together all hypergeometric tests for CRISPRi
  crispri.hyper.tests <- do.call(rbind, crispri.hyper.tests.per.dose)
  crispri.osr.enrichment.test.list[[d]] <- crispri.hyper.tests
}

crispri.osr.enrichment.tests <- do.call(rbind, crispri.osr.enrichment.test.list)
# crispri.enrichment.tests$adj.p.value <- p.adjust(crispri.enrichment.tests$hyper.test.p.value, method = 'fdr')
crispri.osr.enrichment.tests.filtered <- crispri.osr.enrichment.tests %>% filter(hyper.test.p.value <= 0.05)
# crispri.osr.enrichment.tests.filtered$overlapping.genes <- sapply(crispri.osr.enrichment.tests.filtered$overlapping.genes, function(x) paste(x, collapse = ', '))
```

```{r}
write_csv(crispri.osr.enrichment.tests.filtered, file = '../data/Fig7/osr-network-enrichment-in-crispri.csv')
```

```{r}
# reading in enrichment file generating in previous script
crispri.enrichment.tests.filtered <- read_csv(file = '../data/Fig6/crispri-enrichment-tests-filtered.csv')
```

```{r}
# generating Table 2, using information from the significant BP genes enriched in CRISPRi and the OSR network enriched in CRISPRi
bayes.v.crispri <- crispri.enrichment.tests.filtered
osr.v.crispri <- crispri.osr.enrichment.tests.filtered
# summarize BP vs. CRISPRi
bayes.v.crispri.summary.list <- list()
for (r in 1:nrow(bayes.v.crispri)) {
  select.test <- bayes.v.crispri[r, ]
  gene.list.conversion <- tibble(rv.number = unlist(select.test$overlapping.genes))
  gene.list.conversion <- suppressMessages(left_join(gene.list.conversion, gene.name.conversion.table))
  bayes.v.crispri.summary.list[[r]] <- tibble(`BP vs. CRISPRi hypergeometric test p value` = select.test$hyper.test.p.value,
                                              `Sig. BP genes` = list(gene.list.conversion$locus),
                                              `CRISPRi treatment` = paste(select.test$response, ' in ', select.test$treatment, sep = ''))
}
bayes.v.crispri.summary <- do.call(rbind, bayes.v.crispri.summary.list)

# summarize OSR vs. CRISPRi
osr.v.crispri.summary.list <- list()
for (r in 1:nrow(osr.v.crispri)) {
  select.test <- osr.v.crispri[r, ]
  gene.list.conversion <- tibble(rv.number = unlist(select.test$overlapping.genes))
  gene.list.conversion <- suppressMessages(left_join(gene.list.conversion, gene.name.conversion.table))
  osr.v.crispri.summary.list[[r]] <- tibble(`OSR vs. CRISPRi hypergeometric test p value` = select.test$hyper.test.p.value,
                                            `OSR genes` = list(gene.list.conversion$locus),
                                            `CRISPRi treatment` = paste(select.test$response, ' in ', select.test$treatment, sep = ''))
}
osr.v.crispri.summary <- do.call(rbind, osr.v.crispri.summary.list)

# collecting treatment conditions in common between the two sets of enrichment tests
common.treatments <- intersect(x = bayes.v.crispri.summary$`CRISPRi treatment`, y = osr.v.crispri.summary$`CRISPRi treatment`)
common.test.tables <- list()
for (c in 1:length(common.treatments)) {
  common.bayes.test <- bayes.v.crispri.summary %>% filter(`CRISPRi treatment` == common.treatments[c])
  common.osr.test <- osr.v.crispri.summary %>% filter(`CRISPRi treatment` == common.treatments[c])
  three.way.genes <- intersect(x = unlist(common.bayes.test$`Sig. BP genes`),
                               y = unlist(common.osr.test$`OSR genes`))
  
  common.test.tables[[c]] <- tibble(`CRISPRi treatment` = common.treatments[c],
                                    `BP vs. CRISPRi hypergeometric test p value` = common.bayes.test$`BP vs. CRISPRi hypergeometric test p value`,
                                    `Sig. BP genes` = common.bayes.test$`Sig. BP genes`,
                                    `OSR vs. CRISPRi hypergeometric test p value` = common.osr.test$`OSR vs. CRISPRi hypergeometric test p value`,
                                    `OSR genes` = common.osr.test$`OSR genes`,
                                    `Three way overlapping genes` = list(three.way.genes))
}
common.tests <- do.call(rbind, common.test.tables)

# collecting significant tests unique to either the BP vs. CRISPRi or OSR vs. CRISPRi results
bayes.v.crispri.unique <- bayes.v.crispri.summary %>% filter(!`CRISPRi treatment` %in% common.treatments)
osr.v.crispri.unique <- osr.v.crispri.summary %>% filter(!`CRISPRi treatment` %in% common.treatments)

all.tests.summary <- bind_rows(common.tests, bayes.v.crispri.unique, osr.v.crispri.unique)

# splitting columns with lists of genes for exporting
all.tests.summary$`Sig. BP genes` <- sapply(all.tests.summary$`Sig. BP genes`, function(x) paste(x, collapse = ', '))
all.tests.summary$`OSR genes` <- sapply(all.tests.summary$`OSR genes`, function(x) paste(x, collapse = ', '))
all.tests.summary$`Three way overlapping genes` <- sapply(all.tests.summary$`Three way overlapping genes`, function(x) paste(x, collapse = ', '))

# further annotating table with its own information
all.tests.summary$`Gene Group` <- unlist(lapply(strsplit(all.tests.summary$`CRISPRi treatment`, ' in'), '[', 1))
all.tests.summary$Dose <- unlist(lapply(strsplit(unlist(lapply(strsplit(all.tests.summary$`CRISPRi treatment`, 'in '), '[', 2)), ' dose'), '[', 1))
all.tests.summary$Antibiotic <- unlist(lapply(strsplit(unlist(lapply(strsplit(all.tests.summary$`CRISPRi treatment`, 'dose '), '[', 2)), ' '), '[', 1))
all.tests.summary$`Day(s)` <- as.numeric(unlist(lapply(strsplit(unlist(lapply(strsplit(all.tests.summary$`CRISPRi treatment`, ' '), '[', 6)), 'd'), '[', 1)))

# factoring and reordering
all.tests.summary$Antibiotic <- factor(all.tests.summary$Antibiotic, levels = c('INH', 'EMB', 'CLR', 'RIF', 'STR', 'BDQ', 'LZD', 'VAN'))
all.tests.summary$Dose <- factor(all.tests.summary$Dose, levels = c('low', 'medium', 'high'))
all.tests.summary$`Gene Group` <- factor(all.tests.summary$`Gene Group`, levels = c('Enriched', 'Depleted', 'Responder'))

all.tests.summary <- all.tests.summary[with(all.tests.summary, order(Antibiotic, Dose, `Day(s)`, `Gene Group`)), ]
all.tests.summary <- all.tests.summary %>% dplyr::select(Antibiotic, Dose, `Day(s)`, `Gene Group`,
                                                         `BP vs. CRISPRi hypergeometric test p value`, `Sig. BP genes`,
                                                         `OSR vs. CRISPRi hypergeometric test p value`, `OSR genes`, `Three way overlapping genes`)
```

```{r}
write_csv(all.tests.summary, file = '../supplemental-tables/Supplementary-Data-2-raw.csv')
```
