---
title: "Variant filtering of results from WGS data analysis"
output: html_notebook
---

**Author: Evan Pepper-Tunick**

Last updated: Feb 14, 2025

This notebook is for filtering the variants (mutations) called by any of the three variant callers in our pipeline (VarScan, Samtools, GATK)

WGS data was processed and analyzed using an in house pipeline, which can be found at: https://github.com/baliga-lab/bwa_pipeline

WGS data from this study can be found on the SRA with the following Project ID:

**Important note: to run notebook, user's computer must have ≥ 2.0 GB memory**

```{r}
# importing necessary libraries
{
  library(ggplot2)
  library(ggridges)
  library(tidyverse)
  library(purrr)
  library(drc)
  library(knitr)
  library(factoextra)
  library(naniar)
  library(rstatix)
  library(ggpubr)
  library(scales)
  library(ggrepel)
}

# reading in genome annotation files
gene.map <- read_csv(file = '../data/genome-annotations/msm-mc2155-full-genome-annotation.csv')
rbh.map <- read_csv(file = '../data/genome-annotations/mtb-msm-rbh-map.csv')
```

```{r}
# creating a function for filtering the variants and merging the data with genome annotations
filterVariants <- function(mutation.table, controls) {
  # selecting controls for given experiment
  control.strains <- controls
  # extracting data for controls
  control.snps <- mutation.table %>% filter(ID %in% control.strains)
  # getting all unique variant positions in control strain(s)
  control.snp.positions <- unique(control.snps$POS)
  # removing all control variant positions from experimental data
  exp.snps <- mutation.table %>% filter(!ID %in% control.strains)
  exp.snps.filtered <- exp.snps %>% filter(!POS %in% control.snp.positions)
  exp.snps.filtered$n.callers <- str_count(exp.snps.filtered$VARIANT_CALLERS, ':') + 1
  exp.snps.filtered <- exp.snps.filtered %>% filter(n.callers >= 2)
  # filtering out synonymous snps
  exp.snps.filtered <- exp.snps.filtered %>% filter(EFFECT != 'SYNONYMOUS_CODING')
  exp.snps.filtered$variant.reads <- as.numeric(unlist(lapply(strsplit(exp.snps.filtered$VARIANT_READS, ':'), '[', 2)))
  # removing mutations with less than 20 supporting reads
  exp.snps.filtered <- exp.snps.filtered %>% filter(variant.reads > 20)
  # reformatting and merging with genome annotation data
  names(exp.snps.filtered)[11] <- 'msm.new.locus'
  exp.snps.filtered <- left_join(exp.snps.filtered, gene.map, by = 'msm.new.locus', na_matches = 'never')
  exp.snps.filtered <- left_join(exp.snps.filtered, rbh.map, by = 'msm.locus.tag', na_matches = 'never')
  exp.snps.filtered <- exp.snps.filtered %>% dplyr::select(ID, POS, `1ST_REF_BASE`, ALT, CHANGE, EFFECT, IMPACT, CLASS,
                                                           CODON, AA_CHANGE, SAMTOOLS_FREQ, VARIANT_CALLERS, VARIANT_FREQS,
                                                           VARIANT_READS, FREQ, F1, msm.new.locus, msm.locus.tag, msm.gene, msm.protein.id,
                                                           mtb.locus.tag, mtb.gene.name, fident, alnlen, mismatch, gapopen, qstart, 
                                                           qend, tstart, evalue, bits)
  return(exp.snps.filtered)
}
```

```{r}
# filtering batch one (one-step-selection-batch-1)
control.ids.1 <- c('17_0A', '18_0B', '19_0C', '20_0D')
mutations.1 <- read_csv(file = '../data/wgs-results/one-step-selection-batch-1/all-variant-calls.csv')
mutations.filtered.1 <- filterVariants(mutations.1, control.ids.1)
mutations.filtered.1
```

```{r}
# filtering batch one (one-step-selection-batch-2)
control.ids.2 <- c('WT')
mutations.2 <- read_csv(file = '../data/wgs-results/one-step-selection-batch-2/all-variant-calls.csv')
mutations.filtered.2 <- filterVariants(mutations.2, control.ids.2)
mutations.filtered.2
```

```{r}
# filtering batch one (fluctuation-assay-batch-1)
control.ids.3 <- c('wt-anc1', 'wt-anc2')
mutations.3 <- read_csv(file = '../data/wgs-results/fluctuation-assay-batch-1/all-variant-calls.csv')
mutations.filtered.3 <- filterVariants(mutations.3, control.ids.3)
mutations.filtered.3
```

Finally, collating all mutation data into one file

```{r}
# merging all 3 datasets
all.mutations.filtered <- rbind(mutations.filtered.1, mutations.filtered.2, mutations.filtered.3)
write_csv(all.mutations.filtered, file = '../data/wgs-results/all-isolate-mutations-filtered.csv')
```